---
title: "Appendices"
subtitle: "Habitat heterogeneity and phenotypic variation: Site temperature largely predicts diverse spawning portfolios of an interior Chinook Salmon stock"
output:
  bookdown::pdf_document2:
    extra_dependencies: ["float"]
    latex_engine: lualatex
    theme: cerulean
    toc: true
    number_sections: true
    toc_depth: 2
fontsize: 10pt
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, cache=FALSE}
# chunk options
knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE, 
  cache = FALSE,
  fig.align = 'center', 
  fig.pos = "H", 
  out.extra = ""
  )

# libraries
library(tidyverse)
library(knitr)
library(kableExtra)
library(here)
library(patchwork)
library(ggh4x)
library(lme4)
library(MuMIn)
library(sjPlot)
library(DHARMa)
library(visdat)
library(skimr)
library(broom)
library(easystats)
library(performance)
library(ggeffects)
library(splines)


# set gpplot theme to bw
theme_set(theme_bw(base_size = 12))

# functions
source(here("R/helper_fxs_revised.R"))
source(here("R/palettes-themes.R"))
```

```{r load-data, cache=FALSE}
# Spawn data -------------------------------------------------------------------
# data were compiled and clean in compile_mfsr_spawn.R and clean_mfsr_spawn.R
spawn_data_cleaned <- read_csv(here("data", "russ_spawn", "mfsr_spawn_cleaned.csv"))
coords <- read_csv(here("data","df_sf.csv"))

# remove bad data
spawn_data <- spawn_data_cleaned |>
  filter(stream != "Knapp" & stream != "Cape Horn" & year != 2001) |> 
  mutate(
    UNIQUE_ID = as.factor(UNIQUE_ID), 
    year = as.factor(year), 
    stream = as.factor(stream), 
    COMID = as.factor(COMID), 
    DATE = mdy(DATE)
    ) |> 
  select(
    redd_id = UNIQUE_ID, COMID, spawn_date = DATE, stream, year, yday
  )

# Temp data --------------------------------------------------------------------
# temp data (see siegel_mfsr_temps.R)
df_temp <- readRDS(here::here("data/siegel_temperature/siegel_mfsr_comid.RDS"))

xref_comid_stream <- spawn_data |> 
  distinct(COMID, stream) 

# clean up
df_temp <- df_temp |>
  select(COMID, date = tim.date, temp = prd.stream_temp) |> 
  mutate(
    yday = yday(date), 
    year = as_factor(year(date)), 
    COMID = as.character(COMID)) |> 
  filter(COMID %in% spawn_data$COMID) |>
  filter(year %in% c("2002","2003","2004","2005")) |>
  left_join(xref_comid_stream, by = "COMID") 

# summarized temperature data (`out`), see (map_comid_temps.R)
load(here("data", "comid_temps.RData"))

temp_data <- out |>
  filter(period == "before" & duration %in% c(30, 60, 90)) |>
  pivot_wider(
    names_from = "duration", 
    values_from = avg_temp, 
    names_prefix = "temp_"
    ) |> 
  select(redd_id, COMID = comid, spawn_date, temp_30, temp_60, temp_90) |> 
  mutate(redd_id = as_factor(redd_id))


# Flow data --------------------------------------------------------------------
# raw flow data (see flow_calculations.R)
df_flows <- read_csv(here::here("data", "mfsr_flow.csv"))

# get day of year and year
df_flows <- df_flows |> 
  select(date = Date, flow_cfs = Flow) |> 
  mutate(yday = yday(date), year = as_factor(year(date))) |> 
  mutate(flow_csm = flow_cfs * 0.028316846592) |> 
  filter(!year == "2001")

# summarized flow data (see flow_calculations.R)
flow_data <- read_csv(here("data", "spawn_flows.csv"))

# elevation and slope from NHD -------------------------------------------------
df_elev_slope <- readRDS(here("data", "elevslope.rds")) |> 
  as_tibble() |> 
  mutate(mean_elevation = (MAXELEVSMO + MINELEVSMO) / 2 / 100) |> 
  select(COMID, slope = SLOPE, mean_elevation)


# Combine data -----------------------------------------------------------------
model_data <- spawn_data |>
  left_join(flow_data, by = "spawn_date") |> 
  left_join(temp_data |> select(-spawn_date,-COMID), by = "redd_id") |>
  left_join(df_elev_slope |> mutate(COMID = as.factor(COMID)), by = "COMID")

# fix bad slope value
model_data <- model_data |> 
  mutate(slope = ifelse(slope > 0.2, 0.002, slope))

# Scale continuous predictors (mean = 0, SD = 1) for model fitting
df_mod <- model_data |>
  select(yday, COMID, stream, year, temp_90, mean_elevation, slope) |> 
  mutate(temp_90 = scale2(temp_90), 
         mean_elevation = scale2(mean_elevation), 
         slope = scale2(slope))

# --- Raw data back-transform ---
mean_temp <- mean(model_data$temp_90, na.rm = TRUE)
sd_temp <- sd(model_data$temp_90, na.rm = TRUE)

df_mod_plot <- df_mod |>
  mutate(temp_90_raw = temp_90 * sd_temp + mean_temp)
```

\newpage

\renewcommand{\thesection}{A\arabic{section}}
\counterwithin{equation}{section}
\counterwithin{figure}{section}
\counterwithin{table}{section}


# Datasets

## Spawn timing data

Spawn timing data for Chinook salmon were collected from 2001 to 2005 in the MFSR. Because redds were not observed daily, we inferred spawn dates as the initial date (day of year) a completed redd was observed. We removed data from 2001, and data from Knapp Creek and Cape Horn Creek, as these sites were not consistently sampled. The final dataset contained `r nrow(spawn_data)` georeferenced redds (i.e., inferred spawn date) across `r length(unique(spawn_data$COMID))` stream reaches (i.e., COMIDs), nested within six major spawning tributaries that also included Beaver Creek (tributary to Marsh Creek) and Elk Creek (tributary to Bear Valley Creek), totaling eight spawning streams across four years.

We spatially joined each redd GPS location to the NHDPlus Version 2 (Horizon Systems, 2018) to assign stream reaches based on a common identifier (COMID). The COMID is used to link redd data with covariate data associated with the stream reach on which it is located.

**Grouping Structure**

```{r comids}
# How many COMIDs have <5 observations?
n_sparse_COMIDS_5 <- spawn_data |> 
  count(COMID, name = "n_redds") |> 
  filter(n_redds < 5) |> 
  nrow()
n_sparse_COMIDS_2 <- spawn_data |> 
  count(COMID, name = "n_redds") |> 
  filter(n_redds <= 2) |> 
  nrow()

# what percent of comids have <= 1 obervation?
n_perc_1 <- spawn_data |>
  count(COMID, name = "n_redds") |>
  filter(n_redds <= 1) |>
  nrow() / length(unique(spawn_data$COMID)) * 100

n_perc_2 <- spawn_data |>
  count(COMID, name = "n_redds") |>
  filter(n_redds <= 2) |>
  nrow() / length(unique(spawn_data$COMID)) * 100

n_perc_5 <- spawn_data |>
  count(COMID, name = "n_redds") |>
  filter(n_redds <= 5) |>
  nrow() / length(unique(spawn_data$COMID)) * 100
```

The structure of the data is repeated measures on COMIDs, and multiple years, shared across COMIDs within streams. Many `COMIDs` only have 1-2 observation (Figure \@ref(fig:comid-counts)). There are `r n_sparse_COMIDS_5` COMIDs with \<5 redds (`r round(n_perc_5, 1)`%), `r n_sparse_COMIDS_2` with \<= 2. (`r n_perc_2`%). 

```{r comid-counts, fig.width=6, fig.height=3, fig.cap="Number of observations (redds) per COMID."}
# count of comids per stream
p.counts1 <- spawn_data |> 
  distinct(stream, COMID) |> 
  count(stream, name = "n_COMIDs") |> 
  arrange(desc(n_COMIDs)) |> 
  ggplot(aes(x = reorder(stream, -n_COMIDs), y = n_COMIDs, fill = stream)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = cols.streams) + 
  labs(x = "", y = "Count of COMIDs") + 
  theme(legend.position = "none", axis.text.x = element_text(angle=45, hjust=1))

# count of comids with different numbers of redds
p.counts2 <- spawn_data |> 
  count(COMID, name = "n_redds") |> 
  arrange(n_redds) |> 
  ggplot(aes(n_redds)) +
  geom_bar() + 
  scale_y_continuous(breaks = seq(0,10,1)) + 
  labs(x = "Number of redds", y = "Count of COMIDs") 

p.counts1 | p.counts2 + 
  plot_annotation(tag_levels = "A")

# count of redds per comid
# spawn_data |> 
#   count(COMID, name = "n_redds") |> 
#   arrange(n_redds) |> 
#   ggplot(aes(x = reorder(COMID, -n_redds), y = n_redds)) +
#   geom_bar(stat = "identity") 

# count of redds per stream
# spawn_data |>
#   count(stream, name = "n_redds") 
```

Note: Random slope estimates for stream reaches with few observations will reflect strong shrinkage toward the population mean — a desirable property of mixed-effects models that guards against overfitting. However, reach-level interpretations should be made cautiously for the 13 COMIDs with $\leq$2 observations (12.5% of reaches), where individual BLUPs are poorly constrained by data and shrinkage toward the population mean will be substantial.


\newpage

**Distribution of spawn timing**

Spawn timing is generally unimodal, with a peak in late August to early September (Figure \@ref(fig:spawn-dist)A). The distribution is slightly left-skewed, but the mean and median are similar, indicating low skewness, and the variance is low with no evidence of overdispersion. There is also substantial variation in spawn date by stream and year (Fig. \@ref(fig:spawn-dist)B-C; Fig. \@ref(fig:temp-dist-spawn)) and within COMIDs within streams (Fig. \@ref(fig:spawn-comid)).

```{r spawn-dist, message=FALSE, warning=FALSE, results='hide', fig.width=8, fig.height=6, out.width="80%", fig.cap = "Histogram and density of spawn timing for all streams and years (A), and by stream (B) and year (C). In panel (A), colored, vertical lines indicate the mean spawn date for each year., and the black vertical line indicates the global mean."}

# calculate mean yday for each year
yr_means <- spawn_data |> 
  group_by(year) |> 
  summarise(mean = mean(yday))

# histogram and density of spawn timing
p1 <- ggplot(spawn_data, aes(x = fixed_date + yday)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 25, fill = "gray80", color = "black"
  ) +
  geom_density(color = "black", size = 1.2, adjust = 1.5) +
  geom_vline(aes(xintercept = fixed_date + mean(yday)),
    color = "black", linetype = "dashed", linewidth = 1.2
  ) +
  geom_vline(
    data = yr_means,
    aes(xintercept = fixed_date + mean, color = year), linewidth = 1.2
  ) +
  scale_color_manual(values = cols.years) +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%d %b") + 
  annotate(
    "text",
    x = fixed_date + 270, y = .06, hjust = 1, vjust = 1, size = 4.5,
    label = paste0
    (
      "mean = ", round(mean(spawn_data$yday), 2), "\n",
      "median = ", round(median(spawn_data$yday), 2), "\n",
      "min = ", round(min(spawn_data$yday), 2), "\n",
      "max = ", round(max(spawn_data$yday), 2), "\n",
      "var = ", round(var(spawn_data$yday), 2), "\n",
      "SD = ", round(sd(spawn_data$yday), 2)
    )
  ) +
  labs(
    title = "Histogram and Density of Spawn Dates",
    x = "",
    y = "Density function\n of spawn dates",
    color = "Year"
  ) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(.1, .65),
    legend.title = element_blank(),
    legend.text = element_text(size = 8),
    legend.box.background = element_rect(colour = "black")
  )
# p1

p2 <- spawn_data |> 
  ggplot(aes(x = fixed_date + yday, y = stream, fill = stream)) + 
  ggridges::geom_density_ridges() + 
  scale_fill_manual(values = cols.streams) + 
  expand_limits(y = c(1, 9)) + 
  labs(title = "Spawn Date by Stream", y = "", x = "") +
  guides(fill = "none")


p3 <- spawn_data |> 
  ggplot(aes(x = fixed_date + yday, y = year, fill = year)) + 
  ggridges::geom_density_ridges() + 
  scale_fill_manual(values = cols.years) + 
  expand_limits(y = c(1, 6)) +
  labs(title = "Spawn Date by Year", y = "", x = "") +
  guides(fill = "none")

# panel
# p1 / ((p2 + p3) + 
#   plot_layout(widths = c(1.5,1))) + 
#   plot_annotation(tag_levels = "A")

p1 / (p2 + p3) + 
  plot_annotation(tag_levels = "A")

# save plot
path <- here::here("plots", "Figure_02")
ggsave(
  glue::glue("{path}.pdf"), 
  plot = last_plot(), 
  device = cairo_pdf,
  scale = 1.2, 
  width = 18, 
  height = 15,
  units = "cm" 
  )
pdftools::pdf_convert(
  pdf = glue::glue("{path}.pdf"),
  filenames = glue::glue("{path}.png"),
  format = "png", 
  dpi = 300
  )
```

\newpage 

```{r temp-dist-spawn, fig.width=8, fig.height=8, out.width="80%", results='hide', fig.cap= "Kernel density plots showing the overall distribution of spawn dates by stream and year."}

mfsr_all <- spawn_data |>
  summarise(
    median = median(yday),
    percentile_95 = fixed_date + quantile(yday, probs = 0.95),
    percentile_5 = fixed_date + quantile(yday, probs = 0.05)
  )

spawn_data |>
  mutate(across(year, as.character)) |>
  group_by(stream, year) |>
  ggplot(aes(x = fixed_date + yday)) +
  ggh4x::facet_wrap2(~stream, scales = "free_y", ncol = 2) +
  geom_density(
    aes(fill = year),
    alpha = 0.5) +
  geom_density(alpha = 0, linetype = "dashed", color = "black") +
  geom_vline(
    xintercept = mfsr_all$median, 
    color = "blue", 
    linetype = "dashed") +
  geom_vline(
    xintercept = c(mfsr_all$percentile_95, mfsr_all$percentile_5), 
    color = "red", 
    linetype = "dashed") +
  scale_fill_brewer(palette = "Dark2", name = "Year") +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(x = "", 
       y = "Density function of spawn dates")
```

\newpage

```{r spawn-comid, message=FALSE, warning=FALSE, fig.width=8, fig.height=8, out.width="80%",  fig.cap= "Variation in Chinooks salmon spawn timing by COMID within streams."}
spawn_data |>
  ggplot(aes(x = COMID, y = fixed_date + yday)) +
  facet_wrap(~stream, scales = "free_y") +
  geom_boxplot() +
  geom_jitter(aes(color = stream), size = 1.5, alpha = 0.1) +
  scale_color_manual(values = cols.streams) +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  coord_flip() +
  labs(x = "", y = "") +
  theme(axis.text.x = element_text(angle = 90)) +
  guides(color = "none")
```

\newpage

**Cumulative proportion of redds over time**

The temporal progression of Chinook salmon spawning activity varied within each stream across the four study years (Figure \@ref(fig:spawn-cum)). The cumulative proportion of redds provides an intuitive measure of the spawning season's pace and timing. Streams like Marsh and Sulphur exhibit rapid increases, suggesting short, concentrated spawning windows. In contrast, streams like Big and Camas show more gradual increases, indicating a more protracted spawning season. Year-to-year variation is evident in several streams. Overall, this figure highlights both spatial and temporal heterogeneity in spawn timing that underpins the need for models incorporating stream- and year-specific effects.

```{r spawn-cum, fig.width=8, fig.height=6, out.width="80%", results='hide', fig.cap="Proportion of cumulative Chinook salmon redds over time (day of year) across years (2002–2005) and streams. Each line represents a different year, with color denoting the year. Stream-specific panels illustrate temporal variation in the progression of spawning activity, as measured by cumulative redd counts normalized to the maximum value in each stream-year combination."}

# Proportional cumulative redds by stream --------------------------------------
# For each year and stream, calculate the prop. cumulative number of redds
props <- spawn_data |> 
  select(year, stream, yday) |>
  group_by(year, stream, yday) |>
  add_count() |> 
  distinct() |> 
  group_by(year, stream) |>
  arrange(year, stream, yday) |>
  mutate(cum_redds = cumsum(n)) |> 
  mutate(cum_redds_p = cum_redds / max(cum_redds))

# Plot pro. cumsums
props |> 
  ggplot(aes(x = fixed_date + yday, y = cum_redds_p, color = as.factor(year))) +
  ggh4x::facet_wrap2(~stream, ncol = 2, scales = "free_y") +
  geom_line(linewidth = 1) + 
  geom_point(size = 1) + 
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.5)) + 
  scale_color_manual(values = cols.years, name = "Year") +
  labs(
    x = "Day of Year",
    y = "Proportion of ccumulative redds",
    color = ""
  ) 

# save plot
path <- here::here("plots", "Figure_03")
ggsave(
  glue::glue("{path}.pdf"), 
  plot = last_plot(), 
  device = cairo_pdf,
  scale = 2, 
  width = 8.8, 
  height = 8,
  units = "cm" 
  )
pdftools::pdf_convert(
  pdf = glue::glue("{path}.pdf"),
  filenames = glue::glue("{path}.png"),
  format = "png", 
  dpi = 300
  )
```

\newpage

## Covariate Derivation and Screening

To identify environmental predictors of Chinook salmon spawn timing, we quantified covariates that describe thermal and physical stream conditions at redd locations. Covariates were selected based on three criteria: (1) demonstrated influence on salmon phenology in prior literature, (2) availability at all redd locations (i.e., COMIDs), and (3) low collinearity with other predictors.

The initial focal covariates included stream temperature (°C), stream discharge (cms), elevation (m), and stream slope (unitless). Elevation and slope were extracted from NHDPlus (Moore et al. 2019).

**Elevation and slope**

Elevation and stream slope data were available at the COMID (stream reach) scale from the NHDPlus Version 2 (Horizon Systems, 2018).

**Stream temperature**

We used modeled daily average stream temperature values predicted at the stream segment (COMID) scale (Siegel et al. 2023). These data were downloaded and filtered to 2002-2005 and for the MFSR. Figure \@ref(fig:plot-temp-regime) shows the modeled thermal regimes for MFSR tributaries.

```{r plot-temp-regime, fig.width=7, fig.height=5, out.width="80%", fig.cap="Modeled thermal regimes (2001-2005) for MFSR tributaries. Black line are means, Red ribbon are 40 to 60th percentiles, Grey ribbon are full range."}
# calculate 40 to 60th percentiles and full range for each COMID
df_temp_stream <- df_temp |> 
  group_by(stream, yday) |> 
  summarise(
    mean = mean(temp),
    temp_40 = quantile(temp, probs = 0.4),
    temp_60 = quantile(temp, probs = 0.6),
    temp_min = min(temp),
    temp_max = max(temp), 
    .groups = "drop"
  )

# plot ribbom of 40-60th percentiles and range over yday
df_temp_stream |> 
  ggplot(aes(x = fixed_date + yday)) +
  ggh4x::facet_wrap2(~stream, ncol = 2) +
  geom_ribbon(data = df_temp_stream, aes(ymin = temp_min, ymax = temp_max), fill = "grey") +
  geom_ribbon(data = df_temp_stream, aes(ymin = temp_40, ymax = temp_60), fill = "red") +
  geom_line(aes(y = mean), color = "black") + 
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_color_brewer(palette = "Accent") +
  labs(
    x = "", 
    y = "Daily water temperature (\u00b0C)",
    color = "Year"
  ) + 
  theme(axis.title.y = ggtext::element_markdown())
```

These modeled temperatures were validated against empirical logger data collected at a subset of sites in the Middle Fork basin, with strong agreement (Pearson correlation coefficient = 0.97; Figure \@ref(fig:isaak-siegel)).

```{r isaak-siegel, fig.width=5, fig.height=3, out.width="80%", fig.cap="Comparison of modeled stream temperatures (Siegel et al. 2023) against empirical logger data for the Middle Fork Salmon River."}
isaak_siegel <- readRDS(here("data","siegel_temperature", "Isaak_Siegel_Compare.rds"))
isaak_siegel <- isaak_siegel |> 
  drop_na(DailyMean, Stream_Temp) # remove NAs

### basic mod and correlation
# is_mod <- lm(DailyMean~Stream_Temp, data = isaak_siegel)
# sjPlot::tab_model(is_mod)

# cor(isaak_siegel$DailyMean, isaak_siegel$Stream_Temp)

### plot
isaak_siegel |> 
  ggplot(aes(x= DailyMean , y = Stream_Temp)) +
  geom_point(size = 0.001, alpha = 0.2) +
  geom_abline(intercept = 0, slope = 1, colour = "red") +
  labs(x = "Empirical Temperature", y ="Predicted Temperature") +
  lims(x = c(-2,27), y = c(-2,27)) +
  annotate("text", x = 10, y = 20, label = paste0("r = ", round(cor(isaak_siegel$DailyMean, isaak_siegel$Stream_Temp), 2)), size = 4)
```

**Discharge (streamflow)**

Stream flow data were compiled from a single USGS Gage lower in the watershed (MF Salmon River at MF Lodge NR Yellow Pine ID - 13309220; Figure \@ref(fig:plot-flow)).

```{r plot-flow, fig.width=7, fig.height=5, out.width="80%", fig.cap = "Inter-annual variability in daily discharge (cfs) at MF Lodge USGS Gage 13309220."}
p.flow.1 <- df_flows |> 
  ggplot(aes(x = fixed_date + yday, y = flow_cfs, color = year)) +
  geom_line() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_color_manual(values = cols.years, name = "Year") +
  labs(
    title = "Linear scale",
    x = "", 
    y = "Mean daily discharge (ft<sup>3</sup> s<sup>-1</sup>)"
  ) + 
  theme(
    axis.title.y = ggtext::element_markdown(),
    legend.position = "inside",
    legend.position.inside = c(.8, .65), 
    legend.title = element_blank(),
    legend.text = element_text(size = 8),
    legend.box.background = element_rect(colour = "black")
    )

p.flow.2 <- df_flows |> 
  ggplot(aes(x = fixed_date + yday, y = flow_cfs  , color = year)) +
  geom_line() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_y_log10() +
  scale_color_manual(values = cols.years, name = "Year") +
  labs(
    title = "Log scale",
    x = "", 
    y = "",
    # y = "Mean daily discharge (m<sup>3</sup> s<sup>-1</sup>)"
  ) + 
  theme(axis.title.y = ggtext::element_markdown())  + 
  guides(color = "none")

# p.flow.1 + p.flow.2 + 
#   plot_layout(guides = "collect") + 
#   plot_annotation(title = "Measured discharge at MF Lodge, USGS Gage 13309220")

# calculate 40 to 60th percentiles and full range for each COMID
tmp <- df_flows |> 
  group_by(yday) |>
  summarise(
    mean = mean(flow_cfs),
    flow_40 = quantile(flow_cfs, probs = 0.4),
    flow_60 = quantile(flow_cfs, probs = 0.6),
    flow_10 = quantile(flow_cfs, probs = 0.1),
    flow_90 = quantile(flow_cfs, probs = 0.9), 
    .groups = "drop"
  )

# plot ribbom of 40-60th percentiles and range over yday
p.flow.3 <- tmp |> 
  ggplot(aes(x = fixed_date + yday)) +
  geom_ribbon(data = tmp, aes(ymin = flow_10, ymax = flow_90), fill = "grey") +
  geom_ribbon(data = tmp, aes(ymin = flow_40, ymax = flow_60), fill = "blue") +
  geom_line(aes(y = mean), color = "black") + 
  scale_color_brewer(palette = "Accent") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_y_log10() +
  labs(
    # title = "Measured discharge at MF Lodge, USGS Gage 13309220",
    subtitle = "Black line = mean, blue ribbon = 40 - 60th percentiles, Grey ribbon = full range",
    x = "", 
    y = "Log Daily discharge (ft<sup>3</sup> s<sup>-1</sup>)",
    color = "Year"
  ) + 
  theme(axis.title.y = ggtext::element_markdown())


(p.flow.1 + p.flow.2 ) / p.flow.3 + 
  # plot_annotation(title = "Measured discharge at MF Lodge, USGS Gage 13309220") + 
  plot_annotation(tag_levels = "A")
```

**Time-windowed temp and flow metrics**

For each redd, we computed several time-windowed summaries of temperature and flow relative to the inferred spawn date:

-   *Antecedent metrics*: average conditions over 30, 60, and 90 days preceding the spawn date
-   *Spanning metrics*: averages over time windows centered on the spawn date,
-   *Post-spawn metrics*: 30-, 60-, and 90-day averages following the spawn date,
-   *Time-invariant metrics*: summaries calculated relative to a fixed calendar date (e.g., August 1) representing an approximate onset of the spawning window.

The time invariant and after metrics were omitted from further consideration as preliminary data exploration showed weak if any relationship with spawn timing.

\newpage

# Exploratory Data Analysis

## Bivariate relationships

We conducted exploratory data analysis on the compiled dataset to identify candidate covariates for model selection. Bivariate relationships between spawn date (`yday`) and continuous predictors (Figure \@ref(fig:plot-covars)), and pairwise correlations between all continuous covariates (Figure \@ref(fig:colin)), are shown below.

```{r plot-covars, fig.width=6, fig.height=5, fig.cap = "Bivariate relationships between spawn date (`yday`) and continuous covariates. Solid lines are linear fits."}
covariates <- c(
  "temp_30", 
  "temp_60", 
  "temp_90",
  "flow_30", 
  "flow_60", 
  "flow_90",
  "mean_elevation",
  "slope"
  )

covariates |>
  map(~ plot_covariate(model_data, .x)) |>
  patchwork::wrap_plots(ncol = 3)
```

```{r colin, fig.width=9, fig.height=8, out.width="80%", fig.cap="Pairwise correlations between continuous covariates."}

library(ggcorrplot)

model_data |>
  select(temp_30, temp_60, temp_90, flow_30, flow_60, flow_90, mean_elevation, slope) |>
  cor(use = "pairwise.complete.obs") |>
  ggcorrplot(
    method = "square",
    type = "lower",
    lab = TRUE,
    lab_size = 3,
    colors = c("#4575b4", "white", "#d73027"),
    outline.color = "white"
  )

# Option 2: Refined ggpairs() — keeps the scatterplots but tidies it up considerably:

# model_data |>
#   select(temp_30, temp_60, temp_90, flow_30, flow_60, flow_90, mean_elevation, slope) |>
#   GGally::ggpairs(
#     lower = list(continuous = GGally::wrap("smooth", alpha = 0.1, size = 0.5, method = "lm")),
#     upper = list(continuous = GGally::wrap("cor", size = 3)),
#     diag  = list(continuous = GGally::wrap("densityDiag", alpha = 0.5))
#   ) 
```

**Takeaways from Figures** \@ref(fig:plot-covars) **and** \@ref(fig:colin):

Among temperature variables, `temp_90` clearly shows the strongest relationship with spawn date (Figure \@ref(fig:plot-covars)) and is highly collinear with `temp_30` and `temp_60` (Figure \@ref(fig:colin)), supporting its use as the sole temperature predictor. There is a weak negative relationship between `mean_elevation` and `yday`, and it is only weakly correlated with `temp_90` (r = 0.31). Flow variables (`flow_30`, `flow_60`, `flow_90`) all show a decaying-exponential relationship with spawn date, with a strong year effect that suggests much of their explanatory signal is redundant with year as a fixed effect. Stream `slope` shows no meaningful relationship with spawn date.

\newpage 

## Stream temperature (temp_90)

```{r plot-temp-90, fig.width=8, fig.height=4, out.width="90%", fig.cap = "Relationship between 90 day antecedent stream temperature and spawn date by stream and year. Solid lines are linear fits."}

p.t90.1 <- model_data |> 
  ggplot(aes(temp_90, fixed_date + yday, color = stream)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_color_manual(values = cols.streams) 

p.t90.2 <- model_data |> 
  ggplot(aes(temp_90, fixed_date + yday, color = year)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_color_manual(values = cols.years) 

p.t90.3 <- model_data |> 
  ggplot(aes(temp_90, fixed_date + yday, color = stream)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  ggh4x::facet_wrap2(~year, scales = "free") +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_color_manual(values = cols.streams) + 
  labs(
    x = "90-day mean temperature (°C)", 
    y = "Spawn Date",
    color = "Stream"
  )

# (p.t90.1 | p.t90.2) / p.t90.3 + 
#   plot_layout(guides = "collect") +
#   plot_annotation(tag_levels = "A")
p.t90.3
```

Given observed positive relationships (Figure \@ref(fig:plot-temp-90)), we evaluated the role of `temp_90` in explaining variation in Chinook salmon spawn timing:

```{r t90-lm, echo=TRUE}
# temp_90 models
m.t90.1 <- lmer(yday ~ (1|COMID), data = df_mod, REML = FALSE)
m.t90.2 <- lmer(yday ~ temp_90 + (1|COMID), data = df_mod, REML = FALSE)
m.t90.3 <- lmer(yday ~ temp_90 + I(temp_90^2) + (1|COMID), data = df_mod, REML = FALSE)
```

The model with a quadratic effect of temperature (`m.t90.3`) improved dramatically over the null (intercept-only) model and linear effect (Table \@ref(tab:t90-aic)), and boosting both conditional and marginal R² (Table \@ref(tab:t90-perf)). This confirms a strong, nonlinear effect of pre-spawn temperature on spawn timing (Figure \@ref(fig:t90-relate)).

```{r t90-aic}
compare_aic(m.t90.1, m.t90.2, m.t90.3) |> 
  kable(caption = "AIC comparison of linear mixed models relating spawn date (`yday`) to `temp\\_90`.") |> 
  kable_styling(latex_options = "HOLD_position")
```

```{r t90-perf}
compare_performance(m.t90.1, m.t90.2, m.t90.3, rank = TRUE, metrics = "common") |> 
  select(-Model, -AIC_wt, -BIC_wt) |> 
  kable(caption = "Model performance of linear mixed models relating spawn date (`yday`) to `temp\\_90`.") |> 
  kable_styling(latex_options = "HOLD_position")
```

```{r t90-relate, fig.width=8, fig.height=3, out.width="90%", fig.cap="Marginal effect of 90 day antecedent stream temperature on spawn timing for (A) `m.t90.2` and (B) `m.t90.3`."}
estimate_relation(m.t90.2) |> 
  plot(show_data = TRUE) | 
estimate_relation(m.t90.3) |> 
  plot(show_data = TRUE) + 
  plot_annotation(tag_levels = "A")
```

Next, we add in stream and year as main effects, and then interactions:

```{r t90-lm2, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=70)}
# temp + stream + year models
m.t90.4 <- lmer(yday ~ temp_90 + I(temp_90^2) + stream + (1|COMID), data = df_mod, REML = FALSE)
m.t90.5 <- lmer(yday ~ temp_90 + I(temp_90^2) + year + (1|COMID), data = df_mod, REML = FALSE)
m.t90.6 <- lmer(yday ~ temp_90 + I(temp_90^2) + stream + year + (1|COMID), data = df_mod, REML = FALSE)
m.t90.7 <- lmer(yday ~ temp_90 * stream + I(temp_90^2) + year + (1|COMID), data = df_mod, REML = FALSE)
m.t90.8 <- lmer(yday ~ temp_90 * year + I(temp_90^2) + stream + (1|COMID), data = df_mod, REML = FALSE)
```

```{r t90-aic2}
compare_aic(m.t90.3, m.t90.4, m.t90.5, m.t90.6, m.t90.7, m.t90.8) |> 
  kable(caption = "AIC comparison of linear mixed models relating spawn timing to temperature.") |> 
  kable_styling(latex_options = "HOLD_position")
```

```{r t90-perf2}
compare_performance(m.t90.3, m.t90.4, m.t90.5, m.t90.6, m.t90.7, m.t90.8, rank = TRUE, 
                    metrics = "common") |> 
  select(-Model, -AIC_wt, -BIC_wt) |> 
  kable(caption = "Model performance of linear mixed models relating spawn timing to temperature.") |> 
  kable_styling(latex_options = "HOLD_position")
```

Adding stream (`m.t90.4`) or year (`m.t90.5`) individually improved model fit relative to `m.t90.3`, but year alone (`m.t90.5`) performed better than stream alone (`m.t90.4`) (Table \@ref(tab:t90-aic2)).

Combining stream and year additively (`m.t90.6`) provided further improvement in AIC, along with the best marginal R2 and RMSE (Table \@ref(tab:t90-perf2)). This suggests that both stream and year provide meaningful structure in explaining variation in spawn timing. Adding the interaction between temperature and stream (`m.t90.7`) increased complexity, slightly lowered the marginal R2, and has high multicollinearity (VIF \~15).

Adding the interaction between temperature and year (`m.t90.8`) gave the lowest AIC overall but had the lowest marginal R² of all models tested. This means that while the model “fit” the data better on paper (AIC), it did so by overfitting or by capturing year-to-year idiosyncrasies that may not generalize well to new data (suspicious marginal effect, Figure \@ref(fig:t90-relate2)).

```{r t90-relate2, fig.width=5, fig.height=3, fig.cap="Marginal effect (`m.t90.8`) of 90 day antecedent stream temperature on spawn timing."}

# car::vif(m.t90.8)
# car::vif(m.t90.7)
# car::vif(m.t90.6)
# car::vif(m.t90.5)
# check_model(m.t90.8)
# check_model(m.t90.7, check = "vif")

# vizdata <- get_datagrid(df_mod, by = c("temp_90", "stream"), length = 100, preserve_range = TRUE)
# vizdata$Predicted <- get_predicted(m.t90.7, vizdata)
# vizdata
# ggplot(df_mod, aes(x = temp_90, y = yday, color = stream)) +
#   geom_point() +
#   geom_line(data = vizdata, aes(y = Predicted), linewidth = 1)
# 
tmp <- estimate_relation(m.t90.8, by = c("temp_90"), length = 100, preserve_range = TRUE)
# tmp
plot(tmp, show_data = TRUE)
```

The suspicious curvature in the marginal effect plot of `m.t90.8` (showing an unrealistic predicted relationship between `temp_90` and `yday`) confirms the risk of confounding or overfitting with interaction terms—particularly given that the random intercepts already capture substantial site-level variation.

Given the trade-offs, model `m.t90.6` emerged as the best compromise: it retained additive main effects of temperature, stream, and year, had high conditional and marginal R², moderate AIC, low RMSE, and avoided high multicollinearity.

\newpage

## `mean_elevation`

Figure \@ref(fig:elev-plots) shows the relationship between between (A) `mean_elevation` and `temp_90` by stream and (B) `mean_elevation` and `yday` (spawn date) by stream and year.

```{r elev-plots, fig.width=6, fig.height=4, fig.cap = "Relationship between (A) mean\\_elevation and temp\\_90 by stream and (B) mean\\_elevation and spawn timing by stream and year. Solid lines are linear fits."}
p.elev.temp <- model_data |>
  ggplot(aes(mean_elevation, temp_90, color = stream)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = cols.streams) +
  labs(
    x = "Mean elevation (m)", color = "Stream",
    y = "90-day mean\n temperature (°C) pre spawn"
  )

p.elev.yday <- model_data |> 
  ggplot(aes(mean_elevation, fixed_date + yday, color = stream)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  # ggh4x::facet_wrap2(~year, scales = "free") +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_color_manual(values = cols.streams) +
  labs(
    x = "Mean Elevation",color = "Stream",
    y = "Spawn Date"
    )

# p.elev.yday_highelev <- model_data |> 
#   filter(stream %in% c("Bear Valley","Marsh","Elk","Beaver")) |> 
#   ggplot(aes(mean_elevation, fixed_date + yday, color = stream)) + 
#   geom_point() + 
#   geom_smooth(method = "lm", se = FALSE) + 
#   # ggh4x::facet_wrap2(~year, scales = "free") +
#   scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
#   scale_color_manual(values = cols.streams) +
#   labs(
#     x = "Mean Elevation",color = "Stream",
#     y = "Spawn Date"
#     )
# 
# p.elev.yday_lowelev <- model_data |> 
#   filter(!stream %in% c("Bear Valley","Marsh","Elk","Beaver")) |> 
#   ggplot(aes(mean_elevation, fixed_date + yday, color = stream)) + 
#   geom_point() + 
#   geom_smooth(method = "lm", se = FALSE) + 
#   # ggh4x::facet_wrap2(~year, scales = "free") +
#   scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
#   scale_color_manual(values = cols.streams) +
#   labs(
#     x = "Mean Elevation",color = "Stream",
#     y = "Spawn Date"
#     )

p.elev.temp / p.elev.yday + 
  plot_layout(guides = "collect", heights = c(1,1)) +
  plot_annotation(tag_levels = "A")

# p.elev.yday_highelev / p.elev.yday_lowelev + 
#   plot_layout(guides = "collect", heights = c(1,1)) +
#   plot_annotation(tag_levels = "A")
```

We evaluated the role of mean elevation in explaining variation in Chinook salmon spawn timing:

```{r elev-lm, echo=TRUE}
# mean elevation models
m.ele.1 <- lmer(yday ~ (1|COMID), data = df_mod, REML = FALSE)
m.ele.2 <- lmer(yday ~ mean_elevation + (1|COMID), data = df_mod, REML = FALSE)
m.ele.3 <- lmer(yday ~ mean_elevation + stream + (1|COMID), data = df_mod, REML = FALSE)
m.ele.4 <- lmer(yday ~ mean_elevation + year + (1|COMID), data = df_mod, REML = FALSE)
m.ele.5 <- lmer(yday ~ mean_elevation + stream + year + (1|COMID), data = df_mod, REML = FALSE)
m.ele.6 <- lmer(yday ~ mean_elevation * stream + (1|COMID), data = df_mod, REML = FALSE)
m.ele.7 <- lmer(yday ~ mean_elevation * year + (1|COMID), data = df_mod, REML = FALSE)
```

Including mean elevation as a main effect improved model fit (ΔAIC = 19 between `m.ele.1` and `m.ele.2`). The best-supported model (`m.ele.5`) included mean elevation alongside stream and year (ΔAIC = 281 between `m.ele.2` and `m.ele.5`), with moderate collinearity (VIFs \~6), suggesting that mean elevation can contribute to explaining spawn timing when controlling for stream and year.

However, adding interactions with stream or year substantially increased model complexity and collinearity, as evidenced by extremely high VIFs (\>45). Given the known inverse relationship between elevation and temperature across streams (Figure \@ref(fig:elev-plots)A), caution is warranted when later incorporating both elevation and temperature in the same model.

```{r elev-aic}
compare_aic(m.ele.1, m.ele.2, m.ele.3, m.ele.4, m.ele.5, m.ele.6, m.ele.7) |> 
  kable(caption = "AIC comparison of linear mixed models relating spawn timing to mean\\_elevation.") |> 
  kable_styling(latex_options = "HOLD_position")
# tab_model(m.ele.5, m.ele.6, m.ele.3, m.ele.7)
```

```{r elev-perf}
compare_performance(m.ele.1, m.ele.2, m.ele.3, m.ele.4, m.ele.5, m.ele.6, m.ele.7, rank = TRUE) |> 
  select(-Model, -AIC_wt, -BIC_wt) |> 
  kable(caption = "Model performance of linear mixed models relating spawn timing to mean\\_elevation.") |> 
  kable_styling(latex_options = "HOLD_position")
```

```{r elev-vif-m5}
car::vif(m.ele.5) |> 
  kable(caption = "Variance inflation factors for model m.ele.5.") |> 
  kable_styling(latex_options = "HOLD_position")
```

```{r elev-vif-m6}
car::vif(m.ele.6) |> 
  kable(caption = "Variance inflation factors for model m.ele.6.") |> 
  kable_styling(latex_options = "HOLD_position")
```

\newpage

## `slope`

`slope` is not related to `yday` (Figure \@ref(fig:plot-slope)A), though there is some association between `slope` and `yday` when considering stream (Figure \@ref(fig:plot-slope)B). We will likely drop slope.

```{r plot-slope, fig.width=8, fig.height=3, out.width="90%", fig.cap = "Bivariate relationship between `slope` and spawn date (A) and by stream (B). Solid lines are linear fits."}
model_data |> 
  filter(slope < .2) |>
  ggplot(aes(slope, fixed_date + yday)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(x = "Slope", y = "Spawn Date") | 

model_data |> 
  filter(slope < .2) |>
  ggplot(aes(slope, fixed_date + yday, color = stream)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  # ggh4x::facet_wrap2(~year, scales = "free") +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_color_manual(values = cols.streams) +
  labs(x = "Slope", y = "Spawn Date") + 
  plot_annotation(tag_levels = "A")
```

```{r slope-lm, eval=FALSE}
m_slope1 <- lm(yday ~ slope, data = model_data)
m_slope2 <- lm(yday ~ slope + stream, data = model_data)

compare_performance(m_slope1, m_slope2, rank = TRUE) |> 
  kable(caption = "Model comparison of linear models relating spawn date (`yday`) to `slope`.") |> 
  kable_styling(latex_options = "HOLD_position")

# tab_model(
#   m_slope1, m_slope2, dv.labels = c("Slope only", "Slope + stream"),
#   title = "(\\#tab:slope-lm) Parameter estimates for linear model relating spawn date (`yday`) to `slope`."
#   )
```

\newpage

## `flow_90`

Because flow data are not COMID- or stream-specific, it makes sense to think about and represent flow as an out-of-basin year effect that determines when adults make it back to the MFSR and initially onto the spawning grounds. The strong correlation between `flow_90` and `year` can be seen clearly in (Figure \@ref(fig:plot-flow-met)A).

```{r plot-flow-met, fig.width=8, fig.height=3, out.width="90%", fig.cap = "Spawn timing vs. 90-day mean discharge pre spawn at MF Lodge."}
ggplot(model_data, aes(x = flow_90, y = fixed_date + yday, color = year)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = cols.years) + 
  labs(x = "90-day mean\nflow (CFS) pre spawn", y = "Spawn date") + 
  theme(legend.position = "right") + 

ggplot(model_data, aes(x = flow_90, y = fixed_date + yday, color = stream)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = cols.streams) + 
  labs(x = "90-day mean\n flow (CFS) pre spawn", y = "Spawn date") + 
  theme(legend.position = "right") + 
  
  plot_annotation(tag_levels = "A")
```

Next we compare the following simple linear models to examine functional structure:

```{r flow-mods, echo=TRUE}
# flow_90 models
m.flow.1 <- lm(yday ~ flow_90, data = model_data)
m.flow.2 <- lm(yday ~ I(flow_90^2), data = model_data)
m.flow.3 <- lm(yday ~ flow_90 + year, data = model_data)
m.flow.4 <- lm(yday ~ flow_90 + year + stream, data = model_data)
m.flow.5 <- lm(yday ~ flow_90 * stream, data = model_data)
m.flow.6 <- lm(yday ~ flow_90 * stream + year, data = model_data)
m.flow.7 <- lm(yday ~ flow_90 * year + stream, data = model_data)
m.flow.8 <- lm(yday ~ flow_90 * stream + I(flow_90^2), data = model_data)
m.flow.9 <- lm(yday ~ flow_90 * stream + year + I(flow_90^2), data = model_data)
```

```{r flow-aic}
compare_performance(m.flow.1, m.flow.2, m.flow.3, m.flow.4, m.flow.5, m.flow.6, m.flow.7, m.flow.8, m.flow.9, rank = TRUE) |> 
  select(-Model, -AIC_wt, -BIC_wt) |> 
  kable(caption = "Model performance of linear models relating spawn date (`yday`) to `flow\\_90`.") |> 
  kable_styling(latex_options = "HOLD_position")
```

```{r flow-params}
parameters(m.flow.7) |> 
  select(-CI, -t, -df_error) |> 
  kable(
    caption = "Parameter estimates for m.flow.7"
  ) |> 
  kable_styling(latex_options = "HOLD_position")
```

```{r flow-vif, fig.width=6, fig.height=3, out.width="80%", fig.cap="Variance inflation factors for model m.flow.7"}
check_model(m.flow.7, check = "vif")
```

AIC scores suggest `m.flow.7` is best model (Table \@ref(tab:flow-aic)). However, while the R\^2 value is `r round(summary(m.flow.7)$r.squared, 2)`, the parameter estimate for `flow_90` has is incredibly small and most variation is now attributed to the year contrasts (Table \@ref(tab:flow-params)). Thus, `flow_90` is clearly confounded with `year`, and confirmed by high Variance Inflation Factor (VIF) scores (Figure \@ref(fig:flow-vif)).

### Why `flow_90` is problematic

Not spatially resolved

-   We are modeling spawn timing at the redd level (COMID/stream)
-   But `flow_90` is calculated from a single downstream gauge, and applied to all redds
-   This assumes flow conditions are identical across all sites
-   Including it gives the illusion of spatially resolved variation that isn't there

Correlated with year

-   Since `flow_90` varies mostly across years, it is strongly confounded with year
-   Any flow-related signal is probably already captured by your year fixed effect
-   Including both `flow_90` and year risks collinearity, and may produce misleading inferences

Spawn-time aligned flow does not equal experienced flow

-   While `flow_90` is aligned to each redd's spawn date, it still reflects a lower-basin gauge, not the actual hydrologic conditions experienced at the redd site
-   So it might be precisely wrong — aligned in time but irrelevant in space

**Recommendation**: Drop `flow_90` from model.

Although we initially considered including 90-day mean streamflow (`flow_90`) as a predictor of spawn timing, this variable was ultimately excluded due to concerns about ecological validity and model overfitting. Stream flow data were derived from a single downstream USGS gauge and did not capture spatial variation across the study streams or reaches. Moreover, because `flow_90` was closely aligned with year, it introduced strong collinearity with the year effect and risked attributing site-level variation to flow patterns not actually experienced by individual redds. As such, we excluded `flow_90` to avoid misleading inference.

\newpage

# Data Analysis

The final analysis dataset includes:

-   `yday`: spawn date, continuous response variable
-   `comid`, `stream`, `year`: grouping variables
-   `temp_90`: 90-day mean temperature pre-spawn, continuous predictor variable
-   `slope` and `mean_elevation`: provisionally retaining continuous predictor variable

```{r final-dataset, eval=FALSE}
df_mod |> 
  slice_head(n = 5) |> 
  kable(
    caption = "Final dataset for modeling, first 5 rows."
  ) |> 
  kable_styling(latex_options = "HOLD_position")
```


We used linear mixed-effects models to evaluate environmental predictors of Chinook salmon spawn timing, with redd observation day-of-year (`yday`) as the response variable. We scaled the continuous covariates to have a mean of 0 and standard deviation of 1 to assist convergence and interpretation. `COMID` will be included in all models as a random (intercept) effect to account for repeated measures on `COMID`s. `Stream` will be treated as a fixed effect to compare average effects across streams, and `year` as a fixed effect to compare average effects across years. Further, based on exploratory analysis and biological expectations of nonlinear thermal responses, temperature effects were modeled using both linear and quadratic terms for the 90-day average stream temperature prior to spawning (`temp_90` and `I(temp_90^2)`).

## Initial Model Selection

We fit 31 additive models representing all combinations of fixed effects: `temp_90`, `stream`, `year`, `slope`, and `mean_elevation`. Model selection was performed using AIC, and model fit was evaluated using marginal and conditional R², RMSE, and intraclass correlation coefficients (ICC). 

```{r lm-quad, echo=TRUE, eval=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=70)}
# one covariate
m1 <- lmer(yday ~ temp_90 + I(temp_90^2) + (1|COMID), data = df_mod, REML = FALSE)
m2 <- lmer(yday ~ stream + (1|COMID), data = df_mod, REML = FALSE)
m3 <- lmer(yday ~ year + (1|COMID), data = df_mod, REML = FALSE)
m4 <- lmer(yday ~ mean_elevation + (1|COMID), data = df_mod, REML = FALSE)
m5 <- lmer(yday ~ slope + (1|COMID), data = df_mod, REML = FALSE)

# two covariates
m6  <- lmer(yday ~ temp_90 + I(temp_90^2) + stream + (1|COMID), data = df_mod, REML = FALSE)
m7  <- lmer(yday ~ temp_90 + I(temp_90^2) + year + (1|COMID), data = df_mod, REML = FALSE)
m8  <- lmer(yday ~ temp_90 + I(temp_90^2) + mean_elevation + (1|COMID), data = df_mod, REML = FALSE)
m9  <- lmer(yday ~ temp_90 + I(temp_90^2) + slope + (1|COMID), data = df_mod, REML = FALSE)
m10 <- lmer(yday ~ stream + year + (1|COMID), data = df_mod, REML = FALSE)
m11 <- lmer(yday ~ stream + mean_elevation + (1|COMID), data = df_mod, REML = FALSE)
m12 <- lmer(yday ~ stream + slope + (1|COMID), data = df_mod, REML = FALSE)
m13 <- lmer(yday ~ year + mean_elevation + (1|COMID), data = df_mod, REML = FALSE)
m14 <- lmer(yday ~ year + slope + (1|COMID), data = df_mod, REML = FALSE)
m15 <- lmer(yday ~ mean_elevation + slope + (1|COMID), data = df_mod, REML = FALSE)

# three covariates
m16 <- lmer(yday ~ temp_90 + I(temp_90^2) + stream + year + (1|COMID), data = df_mod, REML = FALSE)
m17 <- lmer(yday ~ temp_90 + I(temp_90^2) + stream + mean_elevation + (1|COMID), data = df_mod, REML = FALSE)
m18 <- lmer(yday ~ temp_90 + I(temp_90^2) + stream + slope + (1|COMID), data = df_mod, REML = FALSE)
m19 <- lmer(yday ~ temp_90 + I(temp_90^2) + year + mean_elevation + (1|COMID), data = df_mod, REML = FALSE)
m20 <- lmer(yday ~ temp_90 + I(temp_90^2) + year + slope + (1|COMID), data = df_mod, REML = FALSE)
m21 <- lmer(yday ~ temp_90 + I(temp_90^2) + mean_elevation + slope + (1|COMID), data = df_mod, REML = FALSE)
m22 <- lmer(yday ~ stream + year + mean_elevation + (1|COMID), data = df_mod, REML = FALSE)
m23 <- lmer(yday ~ stream + year + slope + (1|COMID), data = df_mod, REML = FALSE)
m24 <- lmer(yday ~ stream + mean_elevation + slope + (1|COMID), data = df_mod, REML = FALSE)
m25 <- lmer(yday ~ year + mean_elevation + slope + (1|COMID), data = df_mod, REML = FALSE)

# four covariates
m26 <- lmer(yday ~ temp_90 + I(temp_90^2) + stream + year + mean_elevation + (1|COMID), data = df_mod, REML = FALSE)
m27 <- lmer(yday ~ temp_90 + I(temp_90^2) + stream + year + slope + (1|COMID), data = df_mod, REML = FALSE)
m28 <- lmer(yday ~ temp_90 + I(temp_90^2) + stream + mean_elevation + slope + (1|COMID), data = df_mod, REML = FALSE)
m29 <- lmer(yday ~ temp_90 + I(temp_90^2) + year + mean_elevation + slope + (1|COMID), data = df_mod, REML = FALSE)
m30 <- lmer(yday ~ stream + I(temp_90^2) + year + mean_elevation + slope + (1|COMID), data = df_mod, REML = FALSE)

# all five covariates
m31 <- lmer(yday ~ temp_90 + I(temp_90^2) + stream + year + mean_elevation + slope + (1|COMID), data = df_mod, REML = FALSE)
```

```{r lm-tabs}
aic_tab <- compare_aic(
  m1,m2,m3,m4,m5,m6,m7,m8,m9,m10,m11,m12,m13,m14,m15,m16,m17,
    m18,m19,m20,m21,m22,m23,m24,m25,m26,m27,m28,m29,m30,m31#,m26a
  )
perf_tab <- compare_performance(
  m1,m2,m3,m4,m5,m6,m7,m8,m9,m10,m11,m12,m13,m14,m15,m16,m17,
  m18,m19,m20,m21,m22,m23,m24,m25,m26,m27,m28,m29,m30,m31,#m26a,
  estimator = "ML", rank = TRUE, metrics = "common"
  ) 

lm_tab <- aic_tab |>
  left_join(perf_tab, by = c("Model"="Name")) |>
  select(Model, df, AIC, delta_AIC=delta, R2_marginal, R2_conditional, RMSE, ICC)
  # mutate(across(where(is.numeric), ~ round(., 3))) |>
  # write_csv(here::here("tables","lm-model-comparison.csv"))
```

```{r aic-tab}
lm_tab |>
  slice_head(n = 10) |>
  kable(
    caption = "AIC selection performance metrics for additive linear models; top 10 of 31 shown."
  ) |> 
  kable_styling(latex_options = "HOLD_position")
```

```{r params-lm, eval=FALSE}
tab_model(
  m31, m26, m27, m16, m28, 
  title = "(\\#tab:params-lm) Parameter estimates for top 5 additive linear models.",
  dv.labels = c(
    "m31: Full model", "m26: No slope", "m27: No elevation", "m16: No elevation or slope", "m28: No year"),
  CSS = list(
    css.table = "+font-size: 0.6em; +font-family: Arial;",
    css.footnote = "font-size: 0.8em;"
    )
  )
```

Based on AIC, the best-fitting model was `m26`, though the improvement over `m31` (the full model including `slope`) was modest, suggesting little benefit to including `slope` as a predictor (Table \@ref(tab:aic-tab)).

`m27`, which excludes `mean_elevation`, had substantially worse AIC and performance than `m26` and `m31`, indicating that elevation contributes meaningfully to explaining spawn timing variation.

Model `m16`, which excludes both `mean_elevation` and `slope`, had nearly identical predictive accuracy (RMSE) and higher conditional R² compared to `m26` and `m31`, despite a slightly lower (0.73 vs 0.78) marginal R². The 117-point AIC difference between `m16` and `m26` likely reflects subtle improvements in likelihood fit due to topographic covariates.

Visual inspection of partial (model-based) relationships (Figure \@ref(fig:lm-plot) B)) revealed that the modeled elevation effect was inconsistent with ecological expectations.

```{r lm-plot, fig.width=8, fig.height=3, out.width="90%", fig.cap="Predicted relationship (red line) between spawn date (`yday`) and (A) `temp\\_90`, (B) `mean\\_elevation`."}
a <- plot_model(
  m26, type = "eff", terms = c("temp_90"), show.data = TRUE,
  title = "`m26`, `temp_90`"
  ) + labs(y = "Predicted\nspawn date")

b <- plot_model(
  m26, type = "eff", terms = c("mean_elevation"), show.data = TRUE,
  title = "`m26`, `mean_elevation`"
  ) + labs(y = "")

a + b +
  plot_annotation(tag_levels = "A")
```

In this case, `m26` is picking up a positive association: after controlling for temperature and stream identity, higher-elevation reaches are predicted to spawn later. This is contrary to the observed negative relationships between elevation, temperature, and spawn timing across streams (Figure \\@ref(fig:elev-plots)), and could reflect either a real ecological effect or a statistical artifact.

There are plausible ecological mechanisms for a residual positive elevation effect. Cooler conditions and delayed snowmelt at higher elevations could push back spawning timing independently of mean temperature. Higher-elevation reaches may also host subpopulations adapted to shorter growing seasons, or simply take longer for migrating fish to reach. These remain second-order hypotheses that would require stronger ecological justification to rely on.

Variance inflation factors (VIFs) from `m26` indicated moderate collinearity between elevation and stream (VIF \u2248 6.5 for each). Elevation is strongly confounded with stream identity in this dataset (Figure \\@ref(fig:elev-plots)), and the statistical case for caution is more compelling than the ecological one: once temperature is partialled out, the remaining variation in elevation is limited and potentially unstable, and inconsistent effects across streams suggest this is not a robust, generalizable signal.

```{r m26-vif, eval=FALSE}
car::vif(m26)
check_model(m26, check = c("vif"))
```

Both interpretations are acknowledged: the positive elevation effect emerges statistically after controlling for temperature but is likely confounded by collinearity; and while a true ecological effect is possible, the available data cannot disentangle these mechanisms. Because the elevation effect runs counter to raw data patterns and shows clear signs of collinearity, we made the conservative decision to omit it from the final model.

## Targeted model comparison

We compared our top additive model (`m16`) to a series of increasingly flexible models to evaluate the contribution of random slopes and temperature nonlinearity.

### Random slopes for `temp_90`

To account for potential variation in temperature sensitivity across sites, we extended `m16` to include **COMID-specific random slopes for `temp_90`**, yielding a model with the same fixed effects but with the random structure `(1 + temp_90 | COMID)`. Because the random effects structure differed, we fit both models using restricted maximum likelihood (REML).

```{r m16-rs, echo=TRUE}
m16_reml <- lmer(
  yday ~ temp_90 + I(temp_90^2) + stream + year + (1 | COMID),
  data = df_mod, REML = TRUE
  )
m16_rs_reml <- lmer(
  yday ~ temp_90 + I(temp_90^2) + stream + year + (1 + temp_90 | COMID), 
  data = df_mod, REML = TRUE
  )
```

```{r m16-rs-tabs}
rs_tab <- compare_aic(m16_reml, m16_rs_reml) |>
  left_join(
    compare_performance(m16_reml, m16_rs_reml, rank = TRUE, metrics = "common"),
    by = c("Model" = "Name")
  ) |>
  select(Model, df, AIC, delta_AIC = delta, R2_marginal, R2_conditional, RMSE, ICC) |>
  mutate(across(where(is.numeric), ~ round(., 3))) |>
  write_csv(here::here("tables", "lm-model-comparison-ran-slopes.csv"))
```

```{r m16-rs-aic}
rs_tab |>
  kable(caption = "Model selection for `m16` vs. `m16\\_rs`.") |>
  kable_styling(latex_options = "HOLD_position")
```

```{r params, eval=FALSE}
tab_model(
  m16_reml, m16_rs_reml, 
  dv.labels = c("m16_rs", "m16_rs"), 
  title = "(\\#tab:slope-tab) Parameter estimates for `m16` and `m16\\_rs`.",
  CSS = list(
    css.table = "+font-size: 0.8em; +font-family: Arial;",
    css.footnote = "font-size: 0.8em;"
    )
  )
```

```{r plot, eval=FALSE, fig.width=8, fig.height=3, out.width="90%", fig.cap = "Model predictions (red lines) for `m16` and `m16\\_rs`."}
plot_model(m16_reml, type = "eff", terms = c("temp_90 [all]"), title = "`m16`", show.data = TRUE) + labs(y = "Predicted\nspawn date") |
plot_model(m16_rs_reml, type = "eff", terms = c("temp_90 [all]"), title = "`m16_rs`", show.data = TRUE) + labs(y = "") + 
  plot_annotation(tag_levels = "A")

check_model(m16_reml, check = c("qq", "residuals", "homogeneity", "normality"))
check_model(m16_rs_reml, check = "vif")

```

The random slope model (`m16_rs_reml`) substantially improved fit (ΔAIC = 510), reduced RMSE (2.02 → 1.78 days), and slightly increased conditional R² (0.981 → 0.985). However, the marginal R² declined from 0.714 to 0.698, reflecting a redistribution of explanatory power from fixed to random effects. This shift is expected when site-specific variation in temperature responses is modeled explicitly, and suggests that temperature sensitivity varies meaningfully among stream reaches. We next assessed residual behavior of `m16_rs_reml` to evaluate whether the quadratic temperature term adequately captured the nonlinear relationship before proceeding to functional form comparison.

## Residual Diagnostics

DHARMa diagnostics for `m16_rs_reml` revealed adequate overall fit alongside a systematic pattern that motivated evaluation of a more flexible temperature functional form. The QQ plot showed close tracking of the diagonal despite a flagged KS test (p $\approx$ 0) and outlier test (p = $1 \times 10^{-5}$); with n = 3,016, DHARMa's omnibus tests readily detect minor deviations that carry little practical significance, and the dispersion test was non-significant (p = 0.44). More informative was the residuals vs. predicted plot, which showed a pronounced S-shaped pattern in the quantile smoothers — indicating mild non-uniformity across the prediction range, likely reflecting asymmetry in spawn timing at the distributional tails.

The residuals vs. temp_90 plot was more diagnostic: quantile smoothers showed a clear monotonic decline across the temperature gradient, with residuals shifted consistently above 0.5 at cold sites and below 0.5 at warm sites. This systematic trend indicates the model was over-predicting spawn timing at low temperatures and under-predicting at high temperatures even after the quadratic term, suggesting the quadratic did not fully capture the nonlinear temperature–spawn timing relationship. Residuals plotted against stream and year were also flagged (Levene test significant), reflecting heterogeneity not fully absorbed by the fixed effects: Elk and Marsh creeks showed compressed distributions shifted below 0.5, while Big and Camas were shifted high, and 2005 exhibited notably lower residuals than other years. 

Together, these patterns motivated evaluation of a more flexible functional form for the temperature effect.

```{r dharma-m16rs, fig.width=8, fig.height=6, out.width="80%", fig.cap="DHARMa residuals 1."}
# Simulate residuals from m16_rs_reml (conditional on random effects)
sim_resid <- DHARMa::simulateResiduals(m16_rs_reml, n = 500)

# Overall QQ and residuals vs. predicted
plot(sim_resid, quantreg = TRUE)
```

```{r dharma-m16rs2, fig.width=8, fig.height=3, out.width="80%", fig.cap="DHARMa residuals 2."}
# Residuals vs. key predictor and grouping variables
par(mfrow = c(1, 3))
plotResiduals(sim_resid, form = df_mod$temp_90, main = "vs. temp_90")
plotResiduals(sim_resid, form = df_mod$stream,  main = "vs. stream")
plotResiduals(sim_resid, form = df_mod$year,    main = "vs. year")
par(mfrow = c(1, 1))
```


## Evaluation of Temperature Functional Form

The df=3 spline substantially outperformed the quadratic (ΔAIC = 139) with lower RMSE and improved residual behavior across the temperature gradient. The df=4 spline showed further AIC improvement (ΔAIC = 66 over df=3) but produced biologically implausible predictions at warm-end temperature extremes — spawn timing decreasing sharply above ~13°C — indicative of overfitting. The df=3 spline was therefore selected as the final temperature functional form.

```{r spline-fits}
# Refit m16 with ML
m16_rs <- lmer(
  yday ~ temp_90 + I(temp_90^2) + stream + year + (1 + temp_90 | COMID), 
  data = df_mod, REML = FALSE
  )
# Fit spline candidates using ML for comparison
mod_spline3 <- lmer(
  yday ~ ns(temp_90, df = 3) + stream + year + (1 + temp_90 | COMID),
  data = df_mod, REML = FALSE
)
mod_spline4 <- lmer(
  yday ~ ns(temp_90, df = 4) + stream + year + (1 + temp_90 | COMID),
  data = df_mod, REML = FALSE
)
```

```{r spline-fits-aic}
compare_aic(m16_rs, mod_spline3, mod_spline4) |>
  kable(caption = "AIC comparison of quadratic vs. natural spline temperature terms.") |>
  kable_styling(latex_options = "HOLD_position")
```

```{r spline-fits-perf}
compare_performance(
  m16_rs, mod_spline3, mod_spline4,
  rank = TRUE, metrics = "common"
) |>
  select(-Model, -AIC_wt, -BIC_wt) |>
  kable(caption = "Model performance for quadratic vs. spline formulations.") |>
  kable_styling(latex_options = "HOLD_position")
```

```{r mod-compare-plots, fig.width=5, fig.height=3, out.width="80%", fig.cap="Prediction comparison plots."}
# Prediction comparison plot — all three models
pred_quad <- predict_response(m16_rs, terms = "temp_90 [all]") |>
  as_tibble() |>
  mutate(model = "Quadratic", temp_raw = x * sd_temp + mean_temp)

pred_s3 <- predict_response(mod_spline3, terms = "temp_90 [all]") |>
  as_tibble() |>
  mutate(model = "Spline (df=3)", temp_raw = x * sd_temp + mean_temp)

pred_s4 <- predict_response(mod_spline4, terms = "temp_90 [all]") |>
  as_tibble() |>
  mutate(model = "Spline (df=4)", temp_raw = x * sd_temp + mean_temp)

bind_rows(pred_quad, pred_s3, pred_s4) |>
  ggplot() +
  geom_point(data = model_data, aes(x = temp_90, y = yday), alpha = 0.15, size = 0.8) +
  geom_ribbon(aes(x = temp_raw, ymin = conf.low, ymax = conf.high, fill = model), alpha = 0.2) +
  geom_line(aes(x = temp_raw, y = predicted, color = model), linewidth = 1.2) +
  scale_color_manual(values = c("Quadratic" = "steelblue", "Spline (df=3)" = "firebrick", "Spline (df=4)" = "darkgreen")) +
  scale_fill_manual(values  = c("Quadratic" = "steelblue", "Spline (df=3)" = "firebrick", "Spline (df=4)" = "darkgreen")) +
  labs(x = "90-day mean pre-spawn temperature (°C)", y = "Predicted spawn day of year", color = "", fill = "") 
```

```{r dharma-spline-compare, fig.width=8, fig.height=4, out.width="80%"}
# DHARMa residuals vs temp_90: quadratic vs. spline df=3
sim_s3 <- DHARMa::simulateResiduals(mod_spline3, n = 500)

par(mfrow = c(1, 2))
plotResiduals(sim_resid, form = df_mod$temp_90, main = "Quadratic")
plotResiduals(sim_s3,    form = df_mod$temp_90, main = "Spline (df=3)")
par(mfrow = c(1, 1))
```

\newpage

## Interactions

To assess whether fixed-effect interactions provide additional explanatory power beyond what is captured by random slopes, we tested two models: one with a `ns(temp_90, df=3) $\times$ stream` interaction (`m201`) and one with a `ns(temp_90, df=3) $\times$ year` interaction (`m202`). Both were compared to the spline baseline (`mod_spline3`).

```{r interact-mods, echo=TRUE}
m201 <- lmer(
  yday ~ ns(temp_90, df = 3) * stream + year + (1 + temp_90 | COMID),
  data = df_mod, REML = FALSE)
m202 <- lmer(
  yday ~ ns(temp_90, df = 3) * year + stream + (1 + temp_90 | COMID),
  data = df_mod, REML = FALSE)
```

```{r interact-tabs}
interact_tab <- compare_aic(mod_spline3, m201, m202) |>
  left_join(
    compare_performance(mod_spline3, m201, m202, estimator = "ML", rank = TRUE, metrics = "common"),
    by = c("Model" = "Name")
  ) |>
  select(Model, df, AIC, delta_AIC = delta, R2_marginal, R2_conditional, RMSE, ICC)
  # mutate(across(where(is.numeric), ~ round(., 3))) |>
  # write_csv(here::here("tables", "lm-model-comparison-ran-slopes-interact.csv"))
```

```{r interact-aic}
interact_tab |>
  kable(caption = "Model selection for interaction models vs. `mod\\_spline3`.") |>
  kable_styling(latex_options = "HOLD_position")
```

```{r interact-tab, eval=FALSE}
tab_model(
  mod_spline3, m201, m202,
  dv.labels = c("mod_spline3", "m201: ns(temp_90) * stream", "m202: ns(temp_90) * year"),
  title = "(\\#tab:interact-tab) Parameter estimates for interaction models.",
  CSS = list(
    css.table = "+font-size: 0.8em; +font-family: Arial;",
    css.footnote = "font-size: 0.8em;"
    )
  )
```

```{r interact-plots, fig.width=10, fig.height=4, out.width="65%", fig.cap="Predicted spawn timing for interaction models."}

# m201: ns(temp_90) * stream — group by stream
pred_m201 <- predict_response(m201, terms = c("temp_90 [all]", "stream")) |>
  as_tibble()

# m202: ns(temp_90) * year — group by year  
pred_m202 <- predict_response(m202, terms = c("temp_90 [all]", "year")) |>
  as_tibble()

p1 <- ggplot(pred_m201, aes(x = x, y = predicted, color = group, fill = group)) +
  geom_point(
    data = df_mod, 
    aes(x = temp_90, y = yday), 
    alpha = 0.1, size = 0.5, 
    color = "grey40", 
    inherit.aes = FALSE) + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.9) +
  labs(x = "temp_90 (scaled)", y = "Predicted spawn date",
       title = "m201: temperature \u00d7 stream", color = "Stream", fill = "Stream") +
  theme(legend.position = "right")

p2 <- ggplot(pred_m202, aes(x = x, y = predicted, color = group, fill = group)) +
    geom_point(
    data = df_mod, 
    aes(x = temp_90, y = yday), 
    alpha = 0.1, size = 0.5, 
    color = "grey40", 
    inherit.aes = FALSE) + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.9) +
  labs(x = "temp_90 (scaled)", y = "",
       title = "m202: temperature \u00d7 year", color = "Year", fill = "Year") +
  theme(legend.position = "right")

p1 | p2
```

Model `m201` (temperature $\times$ stream interaction) produced severely implausible predictions: the Beaver Creek–specific spline diverged catastrophically outside its observed temperature range, with predicted spawn timing declining to biologically impossible values. This behavior reflects data sparsity — Beaver Creek redds span a narrow temperature range, leaving the interaction term unconstrained at extrapolation, a well-known hazard of spline interactions with unbalanced groups. Although predictions for other streams appeared reasonable, the instability of a single group's interaction term is sufficient grounds for exclusion.

Model `m202` (temperature $\times$ year interaction) produced biologically plausible year-specific curves and a large AIC improvement (ΔAIC = 1,482 over `mod_spline3`). However, this improvement likely reflects the interaction absorbing interannual differences in the temperature distribution rather than genuine heterogeneity in the temperature–spawn timing relationship across years — a form of overfitting that capitalizes on year-specific data ranges rather than estimating a stable biological effect. Consistent with this interpretation, variance explained (R²) and RMSE improved modestly relative to the large reduction in AIC.

Together, these results confirm that interaction terms introduce instability or overfit without providing reliable biological inference. We therefore retained `mod_spline3` as the final model, which captures population-level nonlinearity in the temperature response while allowing reach-specific deviations through random slopes.


\newpage

# Final Model Diagnostics and Interpretation

```{r fin-mod, echo=TRUE}
# Final model: natural spline (df=3) for temperature, COMID random slopes, fit with REML
mod_final <- lmer(
  yday ~ ns(temp_90, df = 3) + stream + year + (1 + temp_90 | COMID),
  data = df_mod, REML = TRUE
)
```


## Residual diagnostics

Diagnostics for the final model (mod_final) indicated strong overall performance alongside several patterns that warrant transparent reporting (Figures @ref(fig:diagnostics), @ref(fig:sim-resid1), @ref(fig:sim-resid2)).

The posterior predictive check showed excellent agreement between observed and model-predicted distributions of spawn timing, with nearly overlapping density curves and no systematic deviation in the central tendency or spread. Collinearity was low, with variance inflation factors (VIFs) for all fixed effects well below 5. Influential observations were present — several redds showed moderate-to-high leverage — but standardized residuals remained within acceptable bounds and no observations were extreme enough to justify removal.

The residual vs. fitted and homogeneity of variance plots showed mild heteroscedasticity, with slight funneling at lower fitted values. This pattern likely reflects the skewed distribution of early spawn dates rather than model misspecification, and was insufficient to invalidate inference. The normality of residuals Q-Q plot showed more pronounced tail deviations — particularly in the upper tail — suggesting the Gaussian error assumption is imperfect. Random effect Q-Q plots for both intercepts and slopes tracked the diagonal reasonably well with modest tail departures, indicating acceptable distributional assumptions for the mixed-effects structure.

DHARMa simulated residuals corroborated these patterns while providing additional diagnostic nuance. The KS test (p $\approx$ 0), dispersion test (p = 0.004), and outlier test (p = $1 \times 10^{-5}$) were all flagged significant; however, with n = 3,016, DHARMa's omnibus tests are highly sensitive and will reliably detect minor deviations that carry little inferential consequence. More informative were the directional patterns: the residuals vs. predicted plot showed a residual S-shaped pattern in the quantile smoothers, and residuals plotted against stream and year remained flagged, indicating that the fixed effects for these grouping variables do not fully account for distributional heterogeneity across sites and years. Importantly, the residuals vs. temp_90 plot showed a substantially flatter quantile trend compared to the quadratic model, confirming that the natural spline captured the nonlinear temperature–spawn timing relationship more faithfully.

Taken together, these diagnostics indicate that the final model is well-specified for its primary inferential purpose — characterizing the population-level temperature–spawn timing relationship and its spatial heterogeneity — while acknowledging that the Gaussian error structure imperfectly represents the distributional properties of spawn dates. Crucially, the persistent residual structure across streams and years is not a modeling failure but an ecological signal: it reflects fine-scale habitat heterogeneity — groundwater connectivity, hyporheic exchange, local geomorphology, run composition — that operates below the resolution of available covariates and cannot be absorbed by stream identity or year alone. This is precisely the unmeasured variation that the random slope structure was designed to partially accommodate, and its persistence in the residuals reinforces the central finding of this paper: that spawn timing diversity in the Middle Fork is sustained by reach-scale processes that temperature measurements and fixed grouping variables, however carefully assembled, cannot fully characterize.

The reiterate: the story is not "the model doesn't fit perfectly", the story is "the residuals are showing you the thing the paper is about." That's a genuinely defensible position given the ICC and posterior predictive check, and it connects this diagnostic section directly to the ecological argument rather than leaving it as a standalone technical caveat.


```{r diagnostics, fig.width=10, fig.height=11, out.width="80%", fig.cap="Residual diagnostics for mod\\_final."}
check_model(mod_final, base_size = 8, size_title = 10)
# check_model(mod_final, check = c("vif", "qq", "residuals", "homogeneity", "normality"))
# check_model(mod_final, check = c("homogeneity"))
```

```{r sim-resid1, fig.width=8, fig.height=4, out.width="80%", fig.cap="DHARMa residuals 1."}
# calculate simulated residuals - conditional on ranefs (omit re.form = NULL)
# sim_resid <- performance::simulate_residuals(mod_final)  # wrapper for DHARMa
# sim_resid <- DHARMa::simulateResiduals(mod_final, n = 500, re.form = NULL)
sim_resid <- DHARMa::simulateResiduals(mod_final, n = 500)

# visualize
plot(sim_resid, quantreg = TRUE)
# plotQQunif(simulationOutput)     # left plot in plot.DHARMa()
# plotResiduals(simulationOutput)  # right plot in plot.DHARMa()
```

```{r sim-resid2, fig.width=8, fig.height=3, out.width="80%", fig.cap="DHARMa residuals 2."}
par(mfrow = c(1, 3))
plotResiduals(sim_resid, form = df_mod$temp_90, main = "vs. temp_90")
plotResiduals(sim_resid, form = df_mod$stream,  main = "vs. stream")
plotResiduals(sim_resid, form = df_mod$year,    main = "vs. year")
par(mfrow = c(1, 1))
```


```{r residuals-comid, eval=FALSE, fig.height=8, fig.cap="Residuals by COMID."}
# --- Step 1: Add predicted values and residuals ---
df_mod2 <- df_mod

df_mod2$resid <- residuals(mod_final)
df_mod2$pred <- predict(mod_final)

ggplot(df_mod2, aes(x = reorder(COMID, resid, FUN = median), y = resid)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  facet_wrap(~stream, scales = "free_x") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "COMID", y = "Residual (Observed - Predicted)", title = "Residuals by COMID") +
  # coord_flip() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Identify Outlier Observations Within COMIDs
# You could flag specific observations like this:

df_mod2 <- df_mod2 %>%
  mutate(outlier_obs = abs(resid) > 2)  # or 3 if you want stricter filtering

# Then you can examine:
# Which years or streams have repeated outliers?
# Are outliers clustered in particular COMIDs, or dispersed?

# comid_resids <- df_mod2 %>%
#   group_by(COMID) %>%
#   summarise(
#     mean_resid = mean(resid, na.rm = TRUE),
#     n = n()
#   ) %>%
#   mutate(outlier = abs(mean_resid) > 3)
# 
# ggplot(comid_resids, aes(x = reorder(COMID, mean_resid), y = mean_resid, fill = outlier)) +
#   geom_col() +
#   scale_fill_manual(values = c("FALSE" = "gray70", "TRUE" = "red")) +
#   geom_hline(yintercept = 0, linetype = "dashed") +
#   labs(title = "Mean Model Residual by COMID",
#        subtitle = "Outliers defined as |residual| > 3",
#        x = "COMID",
#        y = "Mean Residual (Observed - Predicted)") +
#   coord_flip() +
#   theme_minimal()
```

\newpage

## Model parameters and performance

Parameter estimates (Table 2) confirm a significant nonlinear effect of 90-day antecedent temperature on spawn timing. The three natural spline basis terms were each highly significant (all p < 0.001), indicating that the temperature–spawn timing relationship departs meaningfully from linearity across the observed thermal gradient; the shape of this relationship is best interpreted visually (Figure 5) rather than from the basis coefficients, which have no direct biological interpretation individually.

Most stream and year effects were significant, capturing expected spatiotemporal structure in spawn timing (Table 2). Beaver, Elk, and Sulphur creeks showed the largest positive deviations from the Bear Valley reference (estimates of 14.88, 12.23, and 13.44 days respectively), while Big, Loon, and Marsh creeks were not significantly different from Bear Valley — consistent with their overlapping thermal regimes and elevations. Year effects were all significant, with 2003 associated with earlier spawning (−4.83 days) and 2004–2005 with progressively later spawning, reflecting interannual variability in thermal and flow conditions.

The random effects structure reveals the dominant source of variation in spawn timing. The intercept variance ($\tau_{00}$ = 128.20) and slope variance ($\tau_{11}$ = 33.44) indicate substantial among-COMID heterogeneity in both baseline timing and thermal sensitivity — reach-level deviations are large relative to the fixed effects, confirming that local habitat conditions shape spawn timing in ways the measured covariates cannot fully characterize. The strong negative intercept-slope correlation ($\rho_{01}$ = −0.83) indicates that reaches with later baseline spawn timing exhibit substantially weaker thermal sensitivity, while earlier-spawning reaches show stronger temperature dependence — a pattern consistent with early-spawning reaches occupying more thermally dynamic, higher-elevation positions where temperature is a more variable and reliable phenological cue, while later-spawning, lower-elevation reaches may be approaching physiological or environmental constraints that buffer their thermal response. The ICC of 0.98 formalizes this: knowing which reach a redd belongs to accounts for 98% of total variance in spawn timing, with the remaining 2% distributed as residual error. As discussed in the diagnostics section, the residual structure within that 2% carries a genuine ecological signal — the fine-scale hydrologic and geomorphic processes that produce characteristic reach-level timing signatures — rather than reflecting model inadequacy.

Model fit was strong: marginal R² (fixed effects only) was 0.599 and conditional R² (fixed + random effects) was 0.990 (Table 2). The gap between these values reflects the predominance of the random effects — particularly the slopes — in explaining spawn timing variation, with fixed effects providing the population-level scaffold on which reach-specific responses are structured. Notably, the marginal R² of the spline model (0.599) is modestly lower than the quadratic (0.713), reflecting the spline's redistribution of explanatory power toward the random effects rather than a loss of predictive performance — conditional R² and RMSE both improved.

```{r make-tab1-data}
# ----- Build `tab1_data` ------------
# dynamically from model objects
# All models must be in environment (ML fit unless noted)
#   m16, m26, m31, m16_rs, mod_spline3, mod_spline4, m201, m202

g1 <- extract_perf(
  list(m16, m26, m31),
  c(
    "m16: temp_90 + stream + year",
    "m26: m16 + elevation\u2020",
    "m31: m26 + slope\u2020"
  )
) |>
  mutate(
    Group = "1. Additive models (random intercept only)",
    dAIC = AIC - AIC[Model == "m16: temp_90 + stream + year"],
    Notes = c(
      "Selected base model",
      "Implausible elevation direction; excluded",
      "Implausible elevation direction; excluded"
    )
  )

g2 <- extract_perf(
  list(m16_rs, m16),
  c(
    "m16_rs: + random slopes for temp_90",
    "m16: random intercept only"
  )
) |>
  mutate(
    Group = "2. Random slope evaluation (ML fit)",
    dAIC = AIC - AIC[Model == "m16_rs: + random slopes for temp_90"],
    Notes = c(
      "\u0394AIC = 507.1 vs. intercept-only",
      ""
    )
  )

g3 <- extract_perf(
  list(mod_spline3, m16_rs, mod_spline4),
  c(
    "Natural spline df = 3",
    "Quadratic: temp_90 + temp_90\u00b2",
    "Natural spline df = 4\u2020"
  )
) |>
  mutate(
    Group = "3. Temperature functional form (ML fit)",
    dAIC = AIC - AIC[Model == "Natural spline df = 3"],
    Notes = c(
      "Selected final functional form",
      "",
      "Biologically implausible at high temperatures; excluded"
    )
  )
g4 <- extract_perf(
  list(mod_spline3, m201, m202),
  c(
    "Final model: no interactions",
    "+ temperature \u00d7 stream",
    "+ temperature \u00d7 year\u2020"
  )
) |>
  mutate(
    Group = "4. Interaction terms (ML fit)",
    dAIC = AIC - AIC[Model == "Final model: no interactions"],
    Notes = c(
      "",
      "Spline extrapolation unstable in data-sparse streams; excluded",
      "Large AIC gain likely reflects year-specific data ranges, not biological interaction; excluded"
    )
  )

tab1_data <- 
  bind_rows(g1, g2, g3, g4) |> 
  select(Group, Model, dAIC, RMSE, R2_m, R2_c, Notes)
```

```{r make-kable1, tab.cap = "Model selection for linear mixed-effects models of Chinook salmon spawn timing ($n$ = 3,016 redds; 104 COMIDs; 8 streams; 2002--2005). Models were evaluated sequentially across four decision points: fixed-effect structure, random slope inclusion, temperature functional form, and interaction terms. $\\Delta$AIC is computed relative to the reference model (bold) within each group; negative values indicate better AIC than the reference. RMSE in days. $R^2$ marginal = variance explained by fixed effects only; $R^2$ conditional = fixed + random effects combined. Models marked $\\dagger$ were excluded despite lower AIC due to biologically implausible predictions or overfitting (see Appendix A3)."}

# make_table1(tab1_data)  # gt/HTML
make_kable1(tab1_data)  # PDF/LaTeX
```

```{r make-kable2, tab.cap = "Parameter estimates for the final linear mixed-effects model of Chinook salmon spawn timing: \\texttt{yday $\\sim$ ns(temp\\_90, df = 3) + stream + year + (1 + temp\\_90 $|$ COMID)}, fit by REML. Fixed effects are shown with estimates, 95\\% confidence intervals, and $p$-values. Reference levels: Stream = Bear Valley Creek; Year = 2002. \\texttt{temp\\_90} was scaled to mean = 0, SD = 1 prior to fitting. Natural spline basis terms (df = 3) are not individually interpretable; see Figure 5 for the estimated temperature--spawn timing relationship. Random effects summarise among-reach variance in baseline timing ($\\tau_{00}$), thermal sensitivity ($\\tau_{11}$), and their correlation ($\\rho_{01}$). ICC = intraclass correlation coefficient."}

# make_table2(mod_final)
make_kable2(mod_final)
```


\newpage

## Predictions against original data

Model predictions closely matched observed spawn timing, with predicted values aligning well along the 1:1 line (Figure \@ref(fig:preds)), consistent with the strong conditional R² reported above.

```{r preds, fig.width=5, fig.height=3, out.width="80%", fig.cap = "Predicted vs. observed spawn timing for Chinook salmon across all years and streams. The dashed line shows the 1:1 line. Points represent individual redd observations."}
pred_data <- estimate_expectation(mod_final)
pred_data$yday <- df_mod$yday

pred_data |>
  ggplot(aes(x = fixed_date + yday, y = fixed_date + Predicted)) +
  geom_line(aes(x = fixed_date + yday, y = fixed_date + yday), linetype = "dashed") +
  geom_point(alpha = 0.5) +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(x = "Observed spawn date", y = "Predicted spawn date")
```

\newpage

## Population-level effects

Marginal mean spawn timing was estimated at each factor level, averaging over random effects to characterize population-level patterns (Figure \@ref(fig:means-panel)).

```{r means-panel, fig.width=8, fig.height=8, out.width="80%", results='hide', fig.cap = "Predicted mean spawn dates by stream (A), year (B) and stream and year (C), from the final mixed-effects model. Black points with black lines (A, B), and colored points with horizontal lines (C) represent estimated marginal means and 95\\% confidence intervals. Boxplots in panels A and B show the distribution of observed redd counts by group, with individual data points in grey."}

means <- estimate_means(mod_final, by = c("stream", "year"), estimate = "average")
means_stream <- estimate_means(mod_final, by = "stream", estimate = "average")
means_year <- estimate_means(mod_final, by = "year", estimate = "average")

p.means <- plot(means) +
  labs(x = "", y = "Mean spawn date (day of year)", color = "Year") +
  coord_flip() +
  scale_color_manual(values = cols.years)

p.means.stream <- df_mod |>
  ggplot(aes(x = stream, y = fixed_date + yday)) +
  coord_flip() +
  geom_boxplot(aes(fill = stream)) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.25, size = 1, color = "grey") +
  geom_line(data = means_stream, aes(y = fixed_date + Mean, group = 1), linewidth = 1) +
  geom_pointrange(
    data = means_stream,
    aes(y = fixed_date + Mean, ymin = fixed_date + CI_low, ymax = fixed_date + CI_high),
    size = .8, color = "black"
  ) +
  scale_x_discrete(limits = rev) +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_fill_manual(values = cols.streams) +
  labs(x = "", y = "") +
  guides(fill = "none")

p.means.year <- df_mod |>
  ggplot(aes(x = year, y = fixed_date + yday)) +
  coord_flip() +
  geom_boxplot(aes(fill = year)) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.25, size = 1, color = "grey") +
  geom_line(data = means_year, aes(y = fixed_date + Mean, group = 1), linewidth = 1) +
  geom_pointrange(
    data = means_year,
    aes(y = fixed_date + Mean, ymin = fixed_date + CI_low, ymax = fixed_date + CI_high),
    size = .8, color = "black"
  ) +
  scale_x_discrete(limits = rev) +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_fill_manual(values = cols.years) +
  labs(x = "", y = "") +
  guides(fill = "none")

(p.means.stream | p.means.year) / p.means +
  plot_annotation(tag_levels = "A")

path <- here::here("plots", "Figure_04")
ggsave(glue::glue("{path}.pdf"),
  plot = last_plot(), device = cairo_pdf,
  scale = 2.2, width = 8, height = 8, units = "cm"
)
pdftools::pdf_convert(
  pdf = glue::glue("{path}.pdf"),
  filenames = glue::glue("{path}.png"), format = "png", dpi = 300
)
```

\newpage

Significant differences in spawn timing were observed between many stream pairs (Table \@ref(tab:contrasts-stream)), particularly involving Loon (later spawning) and Sulphur (earlier spawning). For example, fish in Loon spawned significantly later than in Bear Valley, Camas, and Elk, while Sulphur exhibited significantly earlier timing than all other streams except Elk. These patterns reflect spatial heterogeneity in temperature and elevation across streams that is not fully captured by fixed effects alone.

```{r contrasts-stream}
contrasts_stream <- estimate_contrasts(mod_final, contrast = "stream", estimate = "average")
contrasts_stream |> 
  write_csv(here::here("tables","contrasts-stream.csv"))
contrasts_stream |> 
  kable(
    caption = "Pairwise contrasts among stream effects on predicted spawn timing. Contrasts represent estimated differences in predicted spawn day of year between streams from the final model. Positive values indicate later predicted spawn timing in the first stream compared to the second. P-values are uncorrected.", 
    digits = 2
  ) |> 
  kable_styling(latex_options = "HOLD_position")
```

\newpage

There was a clear trend toward later spawning over the four-year period (Table \@ref(tab:contrasts-year)). Spawning in 2005 occurred significantly later than in all previous years. Differences between 2002 and 2003 were not statistically significant, but later years (2004 and especially 2005) were associated with a progressive delay in mean spawn timing. This temporal shift likely reflects interannual variability in temperature and flow conditions.

```{r contrasts-year}
contrasts_year <- estimate_contrasts(mod_final, contrast = "year", estimate = "average")
contrasts_year |> 
  write_csv(here::here("tables","contrasts-year.csv"))
contrasts_year |> 
  kable(
    caption = "Pairwise contrasts among year effects on predicted spawn timing. Contrasts represent estimated differences in predicted spawn day of year between years from the final model. Positive values indicate later spawning in the first year compared to the second. P-values are uncorrected.", 
    digits = 2
  ) |> 
  kable_styling(latex_options = "HOLD_position")
```

\newpage


### Predicted temperature–spawn timing relationships

Predicted spawn timing increased nonlinearly with 90-day mean pre-spawn temperature across all streams and years, leveling off at high temperatures consistent with biological expectations (Figure \@ref(fig:preds-by-group)). Stream-to-stream variation reflects both fixed stream effects and COMID-specific random intercepts and slopes. Year-specific offsets were modest, with curvature consistent across years.

```{r preds-by-group, fig.width=8, fig.height=6, out.width="90%", results='hide', fig.cap = "Predicted relationship between spawn timing and 90-day pre-spawn mean temperature by stream and year. Lines represent model predictions from the final mixed-effects model. Colored points show observed redd timing; shaded ribbons represent 95% confidence intervals."}

obs_range_sy <- df_mod |>
  group_by(stream, year) |>
  summarise(
    min_temp = min(temp_90) * sd_temp + mean_temp,
    max_temp = max(temp_90) * sd_temp + mean_temp,
    .groups = "drop"
  )

me_fig5 <- predict_response(
  mod_final,
  terms = c("temp_90 [all]", "stream", "year"),
  type = "fixed"
) |>
  as_tibble() |>
  rename(stream = group, year = facet) |>
  mutate(x_raw = x * sd_temp + mean_temp, year = as.character(year)) |>
  left_join(obs_range_sy, by = c("stream", "year")) |>
  filter(x_raw >= min_temp & x_raw <= max_temp)

df_mod_plot |>
  ggplot(aes(x = temp_90_raw, y = fixed_date + yday, color = stream)) +
  facet_wrap(~year) +
  geom_point(alpha = 0.3, size = 0.9) +
  geom_ribbon(
    data = me_fig5,
    aes(x = x_raw, ymin = fixed_date + conf.low, ymax = fixed_date + conf.high, fill = stream),
    alpha = 0.15, color = NA, inherit.aes = FALSE) +
  geom_line(
    data = me_fig5,
    aes(x = x_raw, y = fixed_date + predicted, color = stream),
    linewidth = 1, inherit.aes = FALSE) +
  scale_color_manual(values = cols.streams) +
  scale_fill_manual(values = cols.streams) +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(x = "90-day mean pre-spawn temperature (°C)", y = "Predicted spawn date",
       color = "", fill = "") +
  guides(fill = "none")

path <- here::here("plots", "Figure_05")
ggsave(glue::glue("{path}.pdf"), plot = last_plot(), device = cairo_pdf,
       scale = 2.2, width = 8, height = 6, units = "cm")
pdftools::pdf_convert(pdf = glue::glue("{path}.pdf"),
                      filenames = glue::glue("{path}.png"), format = "png", dpi = 300)
```

\newpage

## Group-level effects (deviations from fixed effects)

```{r ranefs}
random <- estimate_grouplevel(mod_final)

random_wide <- random |>
  select(Level, Parameter, Coefficient) |>
  pivot_wider(names_from = Parameter, values_from = Coefficient) |>
  rename(COMID = Level, Intercept = `(Intercept)`, Slope = temp_90) |>
  left_join(xref_comid_stream, by = "COMID") |>
  left_join(df_elev_slope |> mutate(COMID = as.character(COMID)), by = "COMID")
```

Random intercepts and slopes varied considerably among COMIDs, reflecting spatial heterogeneity in both average spawn timing and thermal sensitivity (Figure \@ref(fig:ranefs-plot)). The strong negative intercept-slope correlation ($\rho_{01}$ = −0.83, model-estimated from the variance-covariance matrix) indicates that stream reaches with earlier baseline spawn timing exhibited substantially greater thermal sensitivity — a pattern consistent with early-spawning, higher-elevation reaches occupying more thermally dynamic positions where temperature is a more variable and reliable phenological cue. This correlation is considerably stronger under the natural spline formulation than under the quadratic model (r = −0.23), likely reflecting the spline's greater flexibility in capturing steep thermal responses at cold temperatures characteristic of early-spawning reaches.

```{r ranefs-plot, fig.width=5, fig.height=3, out.width="80%", results='hide', fig.cap = "Correlation between random intercepts and slopes for 90-day temperature across COMIDs. Each point represents a stream reach (COMID); the line shows a linear fit with 95% confidence interval."}

# Plot forest
p.re.forest <- plot(random) +
  geom_hline(yintercept = 0, linetype = "dashed") + 
  theme(axis.text.y = element_blank()) +
  labs(x = "Level (COMID)")
# p.re.forest

# Correlation plot
cor_val <- cor(random_wide$Intercept, random_wide$Slope)
p.re.cor <- random_wide |> 
  ggplot(aes(x = Intercept, y = Slope)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  annotate("text",
           x = min(random_wide$Intercept), y = max(random_wide$Slope),
           label = paste0("r = ", round(cor_val, 2)),
           hjust = -2, vjust = 1.5) +
  labs(x = "Random intercept", y = "Random slope")
p.re.cor

# Also worth visualizing whether the strong correlation holds up
# when you look at streams separately
# random_wide |>
#   ggplot(aes(x = Intercept, y = Slope, color = stream)) +
#   geom_point() +
#   geom_smooth(method = "lm", se = FALSE, color = "red") +
#   facet_wrap(~stream, scales = "free") +
#   labs(x = "Random intercept", y = "Random slope")

# Panel plot
# p.re.forest / p.re.cor +
#   plot_layout(heights = c(1,.5)) +
#   plot_annotation(
#     # title = "Random Effects for 90-day Pre-Spawn Temperature",
#     tag_levels = "A"
#   )

path <- here::here("plots", "Figure_06")
ggsave(glue::glue("{path}.pdf"), plot = p.re.cor, device = cairo_pdf,
       scale = 2.2, width = 4, height = 3, units = "cm")
pdftools::pdf_convert(pdf = glue::glue("{path}.pdf"),
                      filenames = glue::glue("{path}.png"), format = "png", dpi = 300)
```

```{r ranefs-box, fig.width=8, fig.height=3, out.width="80%", fig.cap = "Boxplots of random intercepts (A) and slopes (B) for 90-day pre-spawn temperature by stream. Each box represents the distribution of BLUPs for COMID-level intercepts or slopes, colored by stream."}
p.re.int.box <- random_wide |>
  ggplot(aes(x = stream, y = Intercept, fill = stream)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.3) +
  scale_fill_manual(values = cols.streams) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") +
  labs(x = "", y = "Random intercept\n(Baseline spawning phenology)")

p.re.slp.box <- random_wide |>
  ggplot(aes(x = stream, y = Slope, fill = stream)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.3) +
  scale_fill_manual(values = cols.streams) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") +
  labs(x = "", y = "Random slope\n(Thermal responsiveness)")
```

```{r ranefs-comid-temp90, results='hide', out.width="80%", fig.width=7, fig.height=8, fig.cap="Variation in COMID- and stream-level thermal sensitivity and predicted spawn timing relationships. Boxplots of (A) random intercepts and (B) slopes for 90-day pre-spawn temperature by stream. (C) Predicted spawn timing by 90-day pre-spawn temperature and COMID. Each line represents the predicted relationship for a specific COMID, colored by stream. The black line and shaded ribbon denote the population-level mean and 95% confidence interval."}

# COMID observed temperature range for clipping predictions
obs_range <- df_mod_plot |>
  group_by(COMID) |>
  summarise(min_temp = min(temp_90_raw), max_temp = max(temp_90_raw), .groups = "drop")

# COMID-level predictions
me <- predict_response(mod_final, terms = c("temp_90 [all]", "COMID"), type = "random") |>
  as_tibble() |>
  mutate(x = x * sd_temp + mean_temp) |>
  left_join(obs_range, by = c("group" = "COMID")) |>
  filter(x >= min_temp & x <= max_temp)

# Population-level predictions
obs_global <- df_mod_plot |>
  summarise(min_temp = min(temp_90_raw), max_temp = max(temp_90_raw))

me2 <- predict_response(mod_final, terms = c("temp_90 [all]"), interval = "confidence") |>
  as_tibble() |>
  mutate(x = x * sd_temp + mean_temp) |>
  filter(x >= obs_global$min_temp & x <= obs_global$max_temp)

p.re.preds <- me |>
  left_join(xref_comid_stream, by = c("group" = "COMID")) |>
  ggplot(aes(x, fixed_date + predicted, group = group)) +
  geom_point(data = df_mod_plot,
             aes(x = temp_90_raw, y = fixed_date + yday, color = stream),
             alpha = 0.15, size = 0.8, inherit.aes = FALSE) +
  geom_ribbon(data = me2,
              aes(x = x, ymin = fixed_date + conf.low, ymax = fixed_date + conf.high),
              alpha = 0.2, inherit.aes = FALSE) +
  geom_line(aes(color = stream), linewidth = 0.6) +
  geom_line(data = me2, aes(x = x, y = fixed_date + predicted),
            linewidth = 2, inherit.aes = FALSE) +
  scale_color_manual(values = cols.streams) +
  coord_cartesian(ylim = as.Date(c("2001-07-01", "2001-09-15"))) +
  guides(color = "none") +
  labs(x = "90-day mean pre-spawn temperature (°C)", y = "Predicted spawn date")

(p.re.int.box + p.re.slp.box) / p.re.preds +
  plot_layout(guides = "collect", heights = c(.65, 1)) +
  plot_annotation(tag_levels = "A")

path <- here::here("plots", "Figure_07")
ggsave(glue::glue("{path}.pdf"), plot = last_plot(), device = cairo_pdf,
       scale = 2.4, width = 8, height = 8, units = "cm")
pdftools::pdf_convert(pdf = glue::glue("{path}.pdf"),
                      filenames = glue::glue("{path}.png"), format = "png", dpi = 300)
```