# Final Model Diagnostics and Interpretation

```{r fin-mod, echo=TRUE}
mod_final <- lmer(
  yday ~ ns(temp_90, df = 3) + stream + year + (1 + temp_90 | COMID),
  data = df_mod, REML = TRUE
)
```

## Residual diagnostics

Diagnostics for the final model (mod_final) indicated strong overall performance alongside several patterns that warrant transparent reporting (Figures @ref(fig:diagnostics), @ref(fig:sim-resid1), @ref(fig:sim-resid2)).

The posterior predictive check showed excellent agreement between observed and model-predicted distributions of spawn timing, with nearly overlapping density curves and no systematic deviation in the central tendency or spread. Collinearity was low, with variance inflation factors (VIFs) for all fixed effects well below 5. Influential observations were present — several redds showed moderate-to-high leverage — but standardized residuals remained within acceptable bounds and no observations were extreme enough to justify removal.

The residual vs. fitted and homogeneity of variance plots showed mild heteroscedasticity, with slight funneling at lower fitted values. This pattern likely reflects the skewed distribution of early spawn dates rather than model misspecification, and was insufficient to invalidate inference. The normality of residuals Q-Q plot showed more pronounced tail deviations — particularly in the upper tail — suggesting the Gaussian error assumption is imperfect. Random effect Q-Q plots for both intercepts and slopes tracked the diagonal reasonably well with modest tail departures, indicating acceptable distributional assumptions for the mixed-effects structure.

DHARMa simulated residuals corroborated these patterns while providing additional diagnostic nuance. The KS test (p $\approx$ 0), dispersion test (p = 0.004), and outlier test (p = $1 \times 10^{-5}$) were all flagged significant; however, with n = 3,016, DHARMa's omnibus tests are highly sensitive and will reliably detect minor deviations that carry little inferential consequence. More informative were the directional patterns: the residuals vs. predicted plot showed a residual S-shaped pattern in the quantile smoothers, and residuals plotted against stream and year remained flagged, indicating that the fixed effects for these grouping variables do not fully account for distributional heterogeneity across sites and years. Importantly, the residuals vs. temp_90 plot showed a substantially flatter quantile trend compared to the quadratic model, confirming that the natural spline captured the nonlinear temperature–spawn timing relationship more faithfully.

Taken together, these diagnostics indicate that the final model is well-specified for its primary inferential purpose — characterizing the population-level temperature–spawn timing relationship and its spatial heterogeneity — while acknowledging that the Gaussian error structure imperfectly represents the distributional properties of spawn dates. Crucially, the persistent residual structure across streams and years is not a modeling failure but an ecological signal: it reflects fine-scale habitat heterogeneity — groundwater connectivity, hyporheic exchange, local geomorphology, run composition — that operates below the resolution of available covariates and cannot be absorbed by stream identity or year alone. This is precisely the unmeasured variation that the random slope structure was designed to partially accommodate, and its persistence in the residuals reinforces the central finding of this paper: that spawn timing diversity in the Middle Fork is sustained by reach-scale processes that temperature measurements and fixed grouping variables, however carefully assembled, cannot fully characterize.

The reiterate: the story is not "the model doesn't fit perfectly", the story is "the residuals are showing you the thing the paper is about." That's a genuinely defensible position given the ICC and posterior predictive check, and it connects this diagnostic section directly to the ecological argument rather than leaving it as a standalone technical caveat.

```{r diagnostics, fig.width=10, fig.height=11, out.width="80%", fig.cap="Residual diagnostics for mod\\_final."}
check_model(mod_final, base_size = 8, size_title = 10)
```

```{r sim-resid1, fig.width=8, fig.height=4, out.width="80%", fig.cap="DHARMa residuals 1."}
sim_resid <- DHARMa::simulateResiduals(mod_final, n = 500)
plot(sim_resid, quantreg = TRUE)
```

```{r sim-resid2, fig.width=8, fig.height=3, out.width="80%", fig.cap="DHARMa residuals 2."}
par(mfrow = c(1, 3))
plotResiduals(sim_resid, form = df_mod$temp_90, main = "vs. temp_90")
plotResiduals(sim_resid, form = df_mod$stream,  main = "vs. stream")
plotResiduals(sim_resid, form = df_mod$year,    main = "vs. year")
par(mfrow = c(1, 1))
```

\newpage

## Model parameters and performance

Parameter estimates (Table 2) confirm a significant nonlinear effect of 90-day antecedent temperature on spawn timing. The three natural spline basis terms were each highly significant (all p < 0.001), indicating that the temperature–spawn timing relationship departs meaningfully from linearity across the observed thermal gradient; the shape of this relationship is best interpreted visually (Figure 5) rather than from the basis coefficients, which have no direct biological interpretation individually.

Most stream and year effects were significant, capturing expected spatiotemporal structure in spawn timing (Table 2). Beaver, Elk, and Sulphur creeks showed the largest positive deviations from the Bear Valley reference (estimates of 14.88, 12.23, and 13.44 days respectively), while Big, Loon, and Marsh creeks were not significantly different from Bear Valley — consistent with their overlapping thermal regimes and elevations. Year effects were all significant, with 2003 associated with earlier spawning (−4.83 days) and 2004–2005 with progressively later spawning, reflecting interannual variability in thermal and flow conditions.

The random effects structure reveals the dominant source of variation in spawn timing. The intercept variance ($\tau_{00}$ = 128.20) and slope variance ($\tau_{11}$ = 33.44) indicate substantial among-COMID heterogeneity in both baseline timing and thermal sensitivity — reach-level deviations are large relative to the fixed effects, confirming that local habitat conditions shape spawn timing in ways the measured covariates cannot fully characterize. The strong negative intercept-slope correlation ($\rho_{01}$ = −0.83) indicates that reaches with later baseline spawn timing exhibit substantially weaker thermal sensitivity, while earlier-spawning reaches show stronger temperature dependence — a pattern consistent with early-spawning reaches occupying more thermally dynamic, higher-elevation positions where temperature is a more variable and reliable phenological cue, while later-spawning, lower-elevation reaches may be approaching physiological or environmental constraints that buffer their thermal response. The ICC of 0.98 formalizes this: knowing which reach a redd belongs to accounts for 98% of total variance in spawn timing, with the remaining 2% distributed as residual error. As discussed in the diagnostics section, the residual structure within that 2% carries a genuine ecological signal — the fine-scale hydrologic and geomorphic processes that produce characteristic reach-level timing signatures — rather than reflecting model inadequacy.

Model fit was strong: marginal R² (fixed effects only) was 0.599 and conditional R² (fixed + random effects) was 0.990 (Table 2). The gap between these values reflects the predominance of the random effects — particularly the slopes — in explaining spawn timing variation, with fixed effects providing the population-level scaffold on which reach-specific responses are structured. Notably, the marginal R² of the spline model (0.599) is modestly lower than the quadratic (0.713), reflecting the spline's redistribution of explanatory power toward the random effects rather than a loss of predictive performance — conditional R² and RMSE both improved.

```{r make-tab1-data}
g1 <- extract_perf(
  list(m16, m26, m31),
  c(
    "m16: temp_90 + stream + year",
    "m26: m16 + elevation\u2020",
    "m31: m26 + slope\u2020"
  )
) |>
  mutate(
    Group = "1. Additive models (random intercept only)",
    dAIC  = AIC - AIC[Model == "m16: temp_90 + stream + year"],
    Notes = c(
      "Selected base model",
      "Collinearity and signal captured by random structure; excluded",
      "As m26, negligable improvement; excluded"
    )
  )

g2 <- extract_perf(
  list(m16_rs, m16),
  c(
    "m16_rs: + random slopes for temp_90",
    "m16: random intercept only"
  )
) |>
  mutate(
    Group = "2. Random slope evaluation (ML fit)",
    dAIC  = AIC - AIC[Model == "m16_rs: + random slopes for temp_90"],
    Notes = c("\u0394AIC = 507.1 vs. intercept-only", "")
  )

g3 <- extract_perf(
  list(mod_spline3, m16_rs, mod_spline4),
  c(
    "Natural spline df = 3",
    "Quadratic: temp_90 + temp_90\u00b2",
    "Natural spline df = 4\u2020"
  )
) |>
  mutate(
    Group = "3. Temperature functional form (ML fit)",
    dAIC  = AIC - AIC[Model == "Natural spline df = 3"],
    Notes = c(
      "Selected final functional form",
      "",
      "Biologically implausible at high temperatures; excluded"
    )
  )

g4 <- extract_perf(
  list(mod_spline3, m201, m202),
  c(
    "Final model: no interactions",
    "+ temperature \u00d7 stream",
    "+ temperature \u00d7 year\u2020"
  )
) |>
  mutate(
    Group = "4. Interaction terms (ML fit)",
    dAIC  = AIC - AIC[Model == "Final model: no interactions"],
    Notes = c(
      "",
      "Spline extrapolation unstable in data-sparse streams; excluded",
      "Large AIC gain likely reflects year-specific data ranges, not biological interaction; excluded"
    )
  )

tab1_data <- bind_rows(g1, g2, g3, g4) |>
  select(Group, Model, dAIC, RMSE, R2_m, R2_c, Notes)
```

```{r make-kable1, tab.cap = "Model selection for linear mixed-effects models of Chinook salmon spawn timing ($n$ = 3,016 redds; 104 COMIDs; 8 streams; 2002--2005). Models were evaluated sequentially across four decision points: fixed-effect structure, random slope inclusion, temperature functional form, and interaction terms. $\\Delta$AIC is computed relative to the reference model (bold) within each group; negative values indicate better AIC than the reference. RMSE in days. $R^2$ marginal = variance explained by fixed effects only; $R^2$ conditional = fixed + random effects combined. Models marked $\\dagger$ were excluded despite lower AIC due to biologically implausible predictions or overfitting (see Appendix A3)."}
gtsave(make_table1(tab1_data), filename = here::here("tables", "mod_selection.docx"))
make_kable1(tab1_data)
```

```{r make-kable2, tab.cap = "Parameter estimates for the final linear mixed-effects model of Chinook salmon spawn timing: \\texttt{yday $\\sim$ ns(temp\\_90, df = 3) + stream + year + (1 + temp\\_90 $|$ COMID)}, fit by REML. Fixed effects are shown with estimates, 95\\% confidence intervals, and $p$-values. Reference levels: Stream = Bear Valley Creek; Year = 2002. \\texttt{temp\\_90} was scaled to mean = 0, SD = 1 prior to fitting. Natural spline basis terms (df = 3) are not individually interpretable; see Figure 5 for the estimated temperature--spawn timing relationship. Random effects summarise among-reach variance in baseline timing ($\\tau_{00}$), thermal sensitivity ($\\tau_{11}$), and their correlation ($\\rho_{01}$). ICC = intraclass correlation coefficient."}
gtsave(make_table2(mod_final), filename = here::here("tables", "mod_coefs.docx"))
make_kable2(mod_final)
```

\newpage

## Predictions against original data

Model predictions closely matched observed spawn timing, with predicted values aligning well along the 1:1 line (Figure \@ref(fig:preds)), consistent with the strong conditional R² reported above.

```{r preds, fig.width=5, fig.height=3, out.width="80%", fig.cap = "Predicted vs. observed spawn timing for Chinook salmon across all years and streams. The dashed line shows the 1:1 line. Points represent individual redd observations."}
pred_data <- estimate_expectation(mod_final)
pred_data$yday <- df_mod$yday

pred_data |>
  ggplot(aes(x = fixed_date + yday, y = fixed_date + Predicted)) +
  geom_line(aes(x = fixed_date + yday, y = fixed_date + yday), linetype = "dashed") +
  geom_point(alpha = 0.5) +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(x = "Observed spawn date", y = "Predicted spawn date")
```

\newpage

## Population-level effects

Marginal mean spawn timing was estimated at each factor level, averaging over random effects to characterize population-level patterns (Figure \@ref(fig:means-panel)).

```{r means-panel, fig.width=8, fig.height=8, out.width="80%", results='hide', fig.cap = "Predicted mean spawn dates by stream (A), year (B) and stream and year (C), from the final mixed-effects model. Black points with black lines (A, B), and colored points with horizontal lines (C) represent estimated marginal means and 95\\% confidence intervals. Boxplots in panels A and B show the distribution of observed redd counts by group, with individual data points in grey."}

means        <- estimate_means(mod_final, by = c("stream", "year"), estimate = "average")
means_stream <- estimate_means(mod_final, by = "stream", estimate = "average")
means_year   <- estimate_means(mod_final, by = "year",   estimate = "average")

p.means <- plot(means) +
  labs(x = "", y = "Mean spawn date (day of year)", color = "Year") +
  coord_flip() +
  scale_color_manual(values = cols.years)

p.means.stream <- df_mod |>
  ggplot(aes(x = stream, y = fixed_date + yday)) +
  coord_flip() +
  geom_boxplot(aes(fill = stream)) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.25, size = 1, color = "grey") +
  geom_line(data = means_stream, aes(y = fixed_date + Mean, group = 1), linewidth = 1) +
  geom_pointrange(
    data = means_stream,
    aes(y = fixed_date + Mean, ymin = fixed_date + CI_low, ymax = fixed_date + CI_high),
    size = .8, color = "black"
  ) +
  scale_x_discrete(limits = rev) +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_fill_manual(values = cols.streams) +
  labs(x = "", y = "") +
  guides(fill = "none")

p.means.year <- df_mod |>
  ggplot(aes(x = year, y = fixed_date + yday)) +
  coord_flip() +
  geom_boxplot(aes(fill = year)) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.25, size = 1, color = "grey") +
  geom_line(data = means_year, aes(y = fixed_date + Mean, group = 1), linewidth = 1) +
  geom_pointrange(
    data = means_year,
    aes(y = fixed_date + Mean, ymin = fixed_date + CI_low, ymax = fixed_date + CI_high),
    size = .8, color = "black"
  ) +
  scale_x_discrete(limits = rev) +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_fill_manual(values = cols.years) +
  labs(x = "", y = "") +
  guides(fill = "none")

(p.means.stream | p.means.year) / p.means +
  plot_annotation(tag_levels = "A")

path <- here::here("plots", "Figure_04")
ggsave(glue::glue("{path}.pdf"),
  plot = last_plot(), device = cairo_pdf,
  scale = 2.2, width = 8, height = 8, units = "cm"
)
pdftools::pdf_convert(
  pdf = glue::glue("{path}.pdf"),
  filenames = glue::glue("{path}.png"), format = "png", dpi = 300
)
```

\newpage

Significant differences in spawn timing were observed between many stream pairs (Table \@ref(tab:contrasts-stream)), particularly involving Loon (later spawning) and Sulphur (earlier spawning). For example, fish in Loon spawned significantly later than in Bear Valley, Camas, and Elk, while Sulphur exhibited significantly earlier timing than all other streams except Elk. These patterns reflect spatial heterogeneity in temperature and elevation across streams that is not fully captured by fixed effects alone.

```{r contrasts-stream}
contrasts_stream <- estimate_contrasts(mod_final, contrast = "stream", estimate = "average")
contrasts_stream |>
  write_csv(here::here("tables", "contrasts-stream.csv"))
contrasts_stream |>
  kable(
    caption = "Pairwise contrasts among stream effects on predicted spawn timing. Contrasts represent estimated differences in predicted spawn day of year between streams from the final model. Positive values indicate later predicted spawn timing in the first stream compared to the second. P-values are uncorrected.",
    digits = 2
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

\newpage

There was a clear trend toward later spawning over the four-year period (Table \@ref(tab:contrasts-year)). Spawning in 2005 occurred significantly later than in all previous years. Differences between 2002 and 2003 were not statistically significant, but later years (2004 and especially 2005) were associated with a progressive delay in mean spawn timing. This temporal shift likely reflects interannual variability in temperature and flow conditions.

```{r contrasts-year}
contrasts_year <- estimate_contrasts(mod_final, contrast = "year", estimate = "average")
contrasts_year |>
  write_csv(here::here("tables", "contrasts-year.csv"))
contrasts_year |>
  kable(
    caption = "Pairwise contrasts among year effects on predicted spawn timing. Contrasts represent estimated differences in predicted spawn day of year between years from the final model. Positive values indicate later spawning in the first year compared to the second. P-values are uncorrected.",
    digits = 2
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

\newpage

### Predicted temperature–spawn timing relationships

Predicted spawn timing increased nonlinearly with 90-day mean pre-spawn temperature across all streams and years, leveling off at high temperatures consistent with biological expectations (Figure \@ref(fig:preds-by-group)). Stream-to-stream variation reflects both fixed stream effects and COMID-specific random intercepts and slopes. Year-specific offsets were modest, with curvature consistent across years.

```{r preds-by-group, fig.width=8, fig.height=6, out.width="90%", results='hide', fig.cap = "Predicted relationship between spawn timing and 90-day pre-spawn mean temperature by stream and year. Lines represent model predictions from the final mixed-effects model. Colored points show observed redd timing; shaded ribbons represent 95% confidence intervals."}

obs_range_sy <- df_mod |>
  group_by(stream, year) |>
  summarise(
    min_temp = min(temp_90) * sd_temp + mean_temp,
    max_temp = max(temp_90) * sd_temp + mean_temp,
    .groups = "drop"
  )

me_fig5 <- predict_response(
  mod_final,
  terms = c("temp_90 [all]", "stream", "year"),
  type = "fixed"
) |>
  as_tibble() |>
  rename(stream = group, year = facet) |>
  mutate(x_raw = x * sd_temp + mean_temp, year = as.character(year)) |>
  left_join(obs_range_sy, by = c("stream", "year")) |>
  filter(x_raw >= min_temp & x_raw <= max_temp)

df_mod_plot |>
  ggplot(aes(x = temp_90_raw, y = fixed_date + yday, color = stream)) +
  facet_wrap(~year) +
  geom_point(alpha = 0.3, size = 0.9) +
  geom_ribbon(
    data = me_fig5,
    aes(x = x_raw, ymin = fixed_date + conf.low, ymax = fixed_date + conf.high, fill = stream),
    alpha = 0.15, color = NA, inherit.aes = FALSE) +
  geom_line(
    data = me_fig5,
    aes(x = x_raw, y = fixed_date + predicted, color = stream),
    linewidth = 1, inherit.aes = FALSE) +
  scale_color_manual(values = cols.streams) +
  scale_fill_manual(values = cols.streams) +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(x = "90-day mean pre-spawn temperature (°C)", y = "Predicted spawn date",
       color = "", fill = "") +
  guides(fill = "none")

path <- here::here("plots", "Figure_05")
ggsave(glue::glue("{path}.pdf"), plot = last_plot(), device = cairo_pdf,
       scale = 2.2, width = 8, height = 6, units = "cm")
pdftools::pdf_convert(pdf = glue::glue("{path}.pdf"),
                      filenames = glue::glue("{path}.png"), format = "png", dpi = 300)
```

\newpage

## Group-level effects (deviations from fixed effects)

```{r ranefs}
random <- estimate_grouplevel(mod_final)

random_wide <- random |>
  select(Level, Parameter, Coefficient) |>
  pivot_wider(names_from = Parameter, values_from = Coefficient) |>
  rename(COMID = Level, Intercept = `(Intercept)`, Slope = temp_90) |>
  left_join(xref_comid_stream, by = "COMID") |>
  left_join(df_elev_slope |> mutate(COMID = as.character(COMID)), by = "COMID")
```

Random intercepts and slopes varied considerably among COMIDs, with substantial spread both within and between streams (Figure \@ref(fig:ranefs-panel)A, \@ref(fig:ranefs-panel)B). Random intercepts represent reach-level deviations in baseline spawn timing after accounting for fixed effects: a positive intercept indicates a reach where salmon spawn later than expected given temperature, stream identity, and year; a negative intercept indicates earlier-than-expected spawning. Random slopes represent deviations in thermal sensitivity: steeper slopes indicate stronger temperature dependence, shallower slopes reflect weaker or buffered responses. Among-reach variance was substantial for both baseline timing (τ₀₀ = 128.20) and thermal sensitivity (τ₁₁ = 33.44), and the pattern of variation differed among streams.

The intercept-slope correlation was strongly negative (ρ₀₁ = −0.83; Figure \@ref(fig:ranefs-panel)C), indicating that reaches with earlier baseline spawn timing exhibited substantially greater thermal sensitivity. Early-spawning, likely higher-elevation reaches are not only earlier — they are also more responsive to temperature variation, suggesting greater phenotypic plasticity. In contrast, later-spawning, lower-elevation reaches exhibit shallower thermal responses, possibly reflecting operation near thermal thresholds or stronger constraint by other cues such as photoperiod or migration timing.

Figure \@ref(fig:ranefs-panel)D translates these random effect patterns into predicted temperature–spawn timing relationships for individual COMIDs, illustrating how the combination of varying intercepts and slopes generates a diverse family of reach-specific trajectories around the population mean. While the overall relationship between pre-spawn temperature and spawn timing is positive and nonlinear, the strength and shape of this response varies substantially among reaches, likely reflecting fine-scale differences in hydrology, geomorphology, and local thermal buffering capacity. This reach-level heterogeneity is the mechanism by which fine-scale habitat diversity sustains diverse spawning portfolios within the Middle Fork.

```{r ranefs-plots}
p.re.forest <- plot(random) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme(axis.text.y = element_blank()) +
  labs(x = "Level (COMID)")

p.re.int.box <- random_wide |>
  ggplot(aes(x = stream, y = Intercept, fill = stream)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.3) +
  scale_fill_manual(values = cols.streams) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") +
  labs(x = "", y = "Random intercept\n(Baseline spawning)")

p.re.slp.box <- random_wide |>
  ggplot(aes(x = stream, y = Slope, fill = stream)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.3) +
  scale_fill_manual(values = cols.streams) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = "none") +
  labs(x = "", y = "Random slope\n(Thermal responsiveness)")

rho <- as.data.frame(VarCorr(mod_final))
cor_val_model <- rho$sdcor[rho$var2 == "temp_90" & rho$var1 == "(Intercept)" & !is.na(rho$var2)]

p.re.cor <- random_wide |>
  ggplot(aes(x = Intercept, y = Slope)) +
  geom_point(aes(fill = stream), shape = 21, size = 2.5, color = "black") +
  scale_fill_manual(values = cols.streams) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  annotate(
    "text",
    x = min(random_wide$Intercept),
    y = max(random_wide$Slope),
    label = paste0("rho == ", round(cor_val_model, 2)),
    parse = TRUE,
    hjust = -5.5, vjust = 1.5
  ) +
  labs(x = "Random intercept", y = "Random slope") +
  theme(legend.position = "none")
```

```{r ranefs-plots-temp90}
obs_range <- df_mod_plot |>
  group_by(COMID) |>
  summarise(
    min_temp = min(temp_90_raw),
    max_temp = max(temp_90_raw),
    .groups = "drop"
  )

me <- predict_response(
  mod_final,
  terms = c("temp_90 [all]", "COMID"),
  type = "random"
)

me_filtered <- me |>
  as_tibble() |>
  mutate(x = x * sd_temp + mean_temp) |>
  left_join(obs_range, by = c("group" = "COMID")) |>
  filter(x >= min_temp & x <= max_temp)

obs_global <- df_mod_plot |>
  summarise(
    min_temp = min(temp_90_raw),
    max_temp = max(temp_90_raw)
  )

me2 <- predict_response(
  mod_final,
  terms = c("temp_90 [all]"),
  interval = "confidence"
) |>
  as_tibble() |>
  mutate(x = x * sd_temp + mean_temp) |>
  filter(x >= obs_global$min_temp & x <= obs_global$max_temp)

p.re.preds <- me_filtered |>
  as_tibble() |>
  left_join(xref_comid_stream, by = c("group" = "COMID")) |>
  ggplot(aes(x, fixed_date + predicted, group = group)) +
  geom_point(
    data = df_mod_plot,
    aes(x = temp_90_raw, y = fixed_date + yday, color = stream),
    alpha = 0.15, size = 0.8, inherit.aes = FALSE
  ) +
  geom_line(aes(color = stream), linewidth = 0.6) +
  geom_ribbon(
    data = me2,
    aes(x = x, ymin = fixed_date + conf.low, ymax = fixed_date + conf.high),
    alpha = 0.1, inherit.aes = FALSE
  ) +
  geom_line(
    data = me2,
    aes(x = x, y = fixed_date + predicted),
    linewidth = 2, inherit.aes = FALSE
  ) +
  scale_color_manual(values = cols.streams) +
  coord_cartesian(ylim = as.Date(c("2001-07-01", "2001-09-15"))) +
  guides(color = "none") +
  labs(
    x = "90-day mean temperature pre-spawn (°C)",
    y = "Predicted spawn day of year"
  ) +
  theme_bw()
```

```{r ranefs-panel, fig.width=8, fig.height=9, out.width="90%", results='hide', fig.cap="Reach-level (COMID) variation in random intercepts, random slopes, and predicted spawn timing relationships from the final mixed-effects model. (A) Boxplots of random intercepts by stream, representing reach-level deviations in baseline spawn timing after accounting for fixed effects (temperature, stream identity, and year); positive values indicate later-than-expected spawning, negative values indicate earlier-than-expected spawning. (B) Boxplots of random slopes by stream, representing reach-level deviations in thermal sensitivity (responsiveness of spawn timing to 90-day antecedent temperature); larger positive values indicate stronger temperature dependence. Points in A and B represent individual stream reaches (COMIDs). (C) Correlation between random intercepts and random slopes across all COMIDs, colored by stream. The strong negative relationship (ρ = −0.83) indicates that reaches with earlier baseline spawn timing exhibit substantially greater thermal sensitivity. The line represents a linear fit with 95% confidence interval. (D) Predicted temperature–spawn timing relationships for individual COMIDs (colored lines, colored by stream) and the population mean response (black line with 95% confidence interval). The diversity of reach-specific trajectories around the population mean reflects fine-scale variation in local habitat conditions — including hydrology, geomorphology, and thermal buffering — that structures phenological diversity across the Middle Fork watershed."}

(p.re.int.box + p.re.slp.box) / p.re.cor / p.re.preds +
  plot_layout(guides = "collect", heights = c(.6, .8, 1)) +
  plot_annotation(tag_levels = "A") &
  theme(plot.margin = margin(2, 4, 2, 4))

path <- here::here("plots", "Figure_06")
ggsave(
  glue::glue("{path}.pdf"),
  plot = last_plot(),
  device = cairo_pdf,
  scale = 2.75,
  width = 7,
  height = 9,
  units = "cm"
)
pdftools::pdf_convert(
  pdf = glue::glue("{path}.pdf"),
  filenames = glue::glue("{path}.png"),
  format = "png",
  dpi = 300
)
```
