# Final Model Diagnostics and Interpretation

```{r final-mod-fit, echo=TRUE}
mod_final <- lmer(
  yday ~ ns(temp_90, df = 3) + stream + year + (1 + temp_90 | COMID),
  data = df_mod, REML = TRUE
)
```

## Residual diagnostics

Diagnostics for the final model (`mod_final`) indicated strong overall performance alongside several patterns that warrant transparent reporting (@fig-diagnostics, @fig-sim-resid1), @fig-sim-resid2).

The posterior predictive check showed excellent agreement between observed and model-predicted distributions of spawn timing, with nearly overlapping density curves and no systematic deviation in the central tendency or spread. Collinearity was low, with variance inflation factors (VIFs) for all fixed effects well below 5. Influential observations were present — several redds showed moderate-to-high leverage — but standardized residuals remained within acceptable bounds and no observations were extreme enough to justify removal.

The residual vs. fitted and homogeneity of variance plots showed mild heteroscedasticity, with slight funneling at lower fitted values. This pattern likely reflects the skewed distribution of early spawn dates rather than model misspecification, and was insufficient to invalidate inference. The normality of residuals Q-Q plot showed more pronounced tail deviations — particularly in the upper tail — suggesting the Gaussian error assumption is imperfect. Random effect Q-Q plots for both intercepts and slopes tracked the diagonal reasonably well with modest tail departures, indicating acceptable distributional assumptions for the mixed-effects structure.

DHARMa simulated residuals corroborated these patterns while providing additional diagnostic nuance. The KS test (p $\approx$ 0), dispersion test (p = 0.004), and outlier test (p = $1 \times 10^{-5}$) were all flagged significant; however, with n = 3,016, DHARMa's omnibus tests are highly sensitive and will reliably detect minor deviations that carry little inferential consequence. More informative were the directional patterns: the residuals vs. predicted plot showed a residual S-shaped pattern in the quantile smoothers, and residuals plotted against stream and year remained flagged, indicating that the fixed effects for these grouping variables do not fully account for distributional heterogeneity across sites and years. Importantly, the residuals vs. temp_90 plot showed a substantially flatter quantile trend compared to the quadratic model, confirming that the natural spline captured the nonlinear temperature–spawn timing relationship more faithfully.

Taken together, these diagnostics indicate that the final model is well-specified for its primary inferential purpose — characterizing the population-level temperature–spawn timing relationship and its spatial heterogeneity — while acknowledging that the Gaussian error structure imperfectly represents the distributional properties of spawn dates. Crucially, the persistent residual structure across streams and years is not a modeling failure but an ecological signal: it reflects fine-scale habitat heterogeneity — groundwater connectivity, hyporheic exchange, local geomorphology, run composition — that operates below the resolution of available covariates and cannot be absorbed by stream identity or year alone. This is precisely the unmeasured variation that the random slope structure was designed to partially accommodate, and its persistence in the residuals reinforces the central finding of this paper: that spawn timing diversity in the Middle Fork is sustained by reach-scale processes that temperature measurements and fixed grouping variables, however carefully assembled, cannot fully characterize.

```{r fig-diagnostics, fig.width=10, fig.height=12, out.width="80%", fig.cap="Residual diagnostics for mod\\_final."}
check_model(mod_final, base_size = 8, size_title = 10)
```

```{r fig-sim-resid1, fig.width=8, fig.height=4, out.width="80%", fig.cap="DHARMa residuals 1."}
sim_resid <- DHARMa::simulateResiduals(mod_final, n = 500)
plot(sim_resid, quantreg = TRUE)
```

```{r fig-sim-resid2, fig.width=8, fig.height=3, out.width="80%", fig.cap="DHARMa residuals 2."}
par(mfrow = c(1, 3))
plotResiduals(sim_resid, form = df_mod$temp_90, main = "vs. temp_90")
plotResiduals(sim_resid, form = df_mod$stream,  main = "vs. stream")
plotResiduals(sim_resid, form = df_mod$year,    main = "vs. year")
par(mfrow = c(1, 1))
```

\newpage

## Model parameters and performance

```{r make-tbl1-data}
g1 <- extract_perf(
  list(m7, m5, m4, m1, m6, m3, m2),
  c(
    "m7: temp\u2090 + stream + year",
    "m5: temp\u2090 + year",
    "m4: temp\u2090 + stream",
    "m1: temp\u2090",
    "m6: stream + year",
    "m3: year",
    "m2: stream"
    )
  ) |>
  mutate(
    Group = "1. Additive models (random intercept only)",
    dAIC  = AIC - AIC[Model == "m7: temp\u2090 + stream + year"],
    Notes = c(
      "Selected base model", "", "", "", "", "", ""
    )
  )

g2 <- extract_perf(
  list(m7_rs_reml, m7_reml),
  c(
    "m7_rs: + random slopes for temp\u2090",
    "m7: random intercept only"
  )
) |>
  mutate(
    Group = "2. Random slope evaluation (REML fit)",
    dAIC  = AIC - AIC[Model == "m7_rs: + random slopes for temp\u2090"],
    Notes = c("\u0394AIC = 510 vs. intercept-only", "")
  )

g3 <- extract_perf(
  list(mod_spline3, m7_rs, mod_spline4),
  c(
    "Natural spline df = 3",
    "Quadratic: temp\u2090 + temp\u2090\u00b2",
    "Natural spline df = 4\u2020"
  )
) |>
  mutate(
    Group = "3. Temperature functional form (ML fit)",
    dAIC  = AIC - AIC[Model == "Natural spline df = 3"],
    Notes = c(
      "Selected final functional form",
      "",
      "Biologically implausible at high temperatures; excluded"
    )
  )

tbl1_data <- bind_rows(g1, g2, g3) |>
  select(Group, Model, dAIC, RMSE, R2_m, R2_c, Notes)
```

```{r}
#| label: make-gt-mod-select
#| output: false
#| results: hide

invisible(
  gtsave(
    make_table1(tbl1_data), 
    filename = here::here("tables", "mod_selection.docx")
    )
  )
```

```{r final-aic, tab.cap = "Model selection for linear mixed-effects models of Chinook salmon spawn timing."}
make_kable1(tbl1_data)
```

```{r}
#| label: make-gt-mod-coefs
#| output: false
#| results: hide

invisible(gtsave(
  make_table2(mod_final), 
  filename = here::here("tables", "mod_coefs.docx")
))
```

```{r final-perf, tab.cap = "Parameter estimates for the final linear mixed-effects model of Chinook salmon spawn timing."}
make_kable2(mod_final)
```

\newpage

### Predictions against original data

Model predictions closely matched observed spawn timing, with predicted values aligning well along the 1:1 line (@fig-pred-vs-obs), consistent with the strong conditional R² reported above.

```{r fig-pred-vs-obs, fig.width=5, fig.height=3, out.width="80%", fig.cap = "Predicted vs. observed spawn timing for Chinook salmon across all years and streams. The dashed line shows the 1:1 line. Points represent individual redd observations."}
pred_data <- estimate_expectation(mod_final)
pred_data$yday <- df_mod$yday

pred_data |>
  ggplot(aes(x = fixed_date + yday, y = fixed_date + Predicted)) +
  geom_line(aes(x = fixed_date + yday, y = fixed_date + yday), linetype = "dashed") +
  geom_point(alpha = 0.5) +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(x = "Observed spawn date", y = "Predicted spawn date")
```

\newpage

## Temperature effects

Predicted spawn timing increased nonlinearly with 90-day mean pre-spawn temperature across all streams and years, leveling off at high temperatures consistent with biological expectations (@fig-temp-effect). Stream-to-stream variation reflects both fixed stream effects and COMID-specific random intercepts and slopes. Year-specific offsets were modest, with curvature consistent across years.

```{r fig-temp-effect, fig.width=8, fig.height=6, out.width="90%", results='hide', fig.cap = "Predicted relationship between spawn timing and 90-day pre-spawn mean temperature by stream and year. Lines represent population-level model predictions from the final mixed-effects model. Colored points illustrate observed spawn timing; shaded ribbons represent 95% confidence intervals."}

obs_range_sy <- df_mod |>
  group_by(stream, year) |>
  summarise(
    min_temp = min(temp_90) * sd_temp + mean_temp,
    max_temp = max(temp_90) * sd_temp + mean_temp,
    .groups = "drop"
  )

me_fig4 <- predict_response(
  mod_final,
  terms = c("temp_90 [all]", "stream", "year"),
  type = "fixed"
) |>
  as_tibble() |>
  rename(stream = group, year = facet) |>
  mutate(x_raw = x * sd_temp + mean_temp, year = as.character(year)) |>
  left_join(obs_range_sy, by = c("stream", "year")) |>
  filter(x_raw >= min_temp & x_raw <= max_temp)

df_mod_plot |>
  ggplot(aes(x = temp_90_raw, y = fixed_date + yday, color = stream)) +
  ggh4x::facet_wrap2(~year, scales = "fixed", axes = "all", strip = strip_vanilla(clip="off")) +
  geom_point(alpha = 0.3, size = 0.9) +
  geom_ribbon(
    data = me_fig4,
    aes(x = x_raw, ymin = fixed_date + conf.low, ymax = fixed_date + conf.high, fill = stream),
    alpha = 0.15, color = NA, inherit.aes = FALSE) +
  geom_line(
    data = me_fig4,
    aes(x = x_raw, y = fixed_date + predicted, color = stream),
    linewidth = 1, inherit.aes = FALSE) +
  coord_cartesian(ylim = as.Date(c("2001-07-01", "2001-09-20"))) +
  scale_color_manual(values = cols.streams) +
  scale_fill_manual(values = cols.streams) +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(x = "90-day mean pre-spawn temperature (°C)", y = "Predicted spawn date",
       color = "", fill = "") +
  guides(fill = "none")

path <- here::here("plots", "Figure_04")
ggsave(glue::glue("{path}.pdf"), plot = last_plot(), device = cairo_pdf,
       scale = 2.2, width = 8, height = 6, units = "cm")
pdftools::pdf_convert(pdf = glue::glue("{path}.pdf"),
                      filenames = glue::glue("{path}.png"), format = "png", dpi = 300)
```

\newpage

## Stream and year effects

Marginal mean spawn timing was estimated at each factor level, averaging over random effects to characterize population-level patterns.

```{r}
#| label: estimate-means-contrasts
#| output: false
#| results: hide

means        <- estimate_means(mod_final, by = c("stream", "year"), estimate = "average")
means_stream <- estimate_means(mod_final, by = "stream", estimate = "average")
means_year   <- estimate_means(mod_final, by = "year",   estimate = "average")

contrasts_stream <- estimate_contrasts(mod_final, contrast = "stream", estimate = "average")
contrasts_stream |> write_csv(here::here("tables", "contrasts-stream.csv"))
contrasts_year <- estimate_contrasts(mod_final, contrast = "year", estimate = "average")
contrasts_year |> write_csv(here::here("tables", "contrasts-year.csv"))
```

```{r print-means, eval=FALSE}
# Network-wide range in predicted spawn timing across streams
cat("=== Stream-level marginal means ===\n")
means_stream |>
  mutate(
    date_mean  = as.Date("2001-01-01") + Mean,
    date_low   = as.Date("2001-01-01") + CI_low,
    date_high  = as.Date("2001-01-01") + CI_high
  ) |>
  select(stream, Mean, date_mean, date_low, date_high) |>
  print()

cat("\n=== Year-level marginal means ===\n")
means_year |>
  mutate(date_mean = as.Date("2001-01-01") + Mean) |>
  select(year, Mean, date_mean, CI_low, CI_high) |>
  print()

cat("\n=== Network summary ===\n")
cat("Earliest stream mean: ", format(as.Date("2001-01-01") + min(means_stream$Mean), "%d %B"), "\n")
cat("Latest stream mean:   ", format(as.Date("2001-01-01") + max(means_stream$Mean), "%d %B"), "\n")
cat("Range (days):         ", round(max(means_stream$Mean) - min(means_stream$Mean), 1), "\n")
cat("Overall network mean: ", format(as.Date("2001-01-01") + mean(means_stream$Mean), "%d %B"), "\n")

```

```{r fig-group-effects, fig.width=8, fig.height=8, out.width="80%", results='hide', fig.cap = "Predicted mean spawn dates by stream (A), year (B), and stream × year (C) from the final mixed-effects model. Black points and lines in panels A and B represent estimated marginal means and 95% confidence intervals averaged over random effects. Colored points show individual redd observations. Colored points and lines in panel C represent stream-specific marginal means and 95% confidence intervals by year."}

# p.means <- plot(means) +
#   labs(x = "", y = "Mean spawn date (day of year)", color = "Year") +
#   coord_flip() +
#   scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
#   scale_color_manual(values = cols.years) + 
#   theme(legend.position = c(0.1, 0.3))

p.means <- means |>
  mutate(
    Mean   = fixed_date + Mean,
    CI_low  = fixed_date + CI_low,
    CI_high = fixed_date + CI_high
  ) |>
  ggplot(aes(x = stream, y = Mean, color = year, group = year)) +
  geom_pointrange(
    aes(ymin = CI_low, ymax = CI_high), 
    position = position_dodge(width = 0.4)
    ) +
  geom_line(position = position_dodge(width = 0.4)) +
  scale_y_date(date_breaks = "1 week", date_labels = "%d %b") +
  scale_color_manual(values = cols.years) +
  # scale_x_discrete(limits = rev) +
  labs(x = "", y = "Mean spawn date", color = "") +
  theme(
    legend.position = c(0.75, 0.4),
    legend.background = element_blank()
  )

p.means.stream <- df_mod |>
  ggplot(aes(x = stream, y = fixed_date + yday, color=stream)) +
  geom_jitter(alpha = 0.5, size = .8) +
  geom_pointrange(
    data = means_stream,
    aes(y = fixed_date + Mean, ymin = fixed_date + CI_low, ymax = fixed_date + CI_high),
    size = .8, color = "black"
  ) +
  # scale_x_discrete(limits = rev) +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_color_manual(values = cols.streams) +
  labs(x = "", y = "Mean spawn date") +
  guides(color = "none") + 
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

p.means.year <- df_mod |>
  ggplot(aes(x = year, y = fixed_date + yday)) +
  geom_jitter(aes(color=year), alpha = 0.5, size = .8) +
  geom_pointrange(
    data = means_year,
    aes(y = fixed_date + Mean, ymin = fixed_date + CI_low, ymax = fixed_date + CI_high),
    size = .8, color = "black"
  ) +
  # scale_x_discrete(limits = rev) +
  scale_y_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_color_manual(values = cols.years) +
  labs(x = "", y = "Mean spawn date") +
  guides(color = "none")

(p.means.stream | p.means.year) / p.means +
  plot_annotation(tag_levels = "A") + 
  plot_layout(heights = c(1,1.5))

path <- here::here("plots", "Figure_05")
ggsave(glue::glue("{path}.pdf"),
  plot = last_plot(), device = cairo_pdf,
  scale = 2.2, width = 8, height = 8, units = "cm"
)
pdftools::pdf_convert(
  pdf = glue::glue("{path}.pdf"),
  filenames = glue::glue("{path}.png"), format = "png", dpi = 300
)
```

\newpage

**Stream contrasts**

Significant differences in spawn timing were observed between many stream pairs (Table A4.3), particularly involving Loon (later spawning) and Sulphur (earlier spawning). For example, fish in Loon spawned significantly later than in Bear Valley, Camas, and Elk, while Sulphur exhibited significantly earlier timing than all other streams except Elk. These patterns reflect spatial heterogeneity in temperature and elevation across streams that is not fully captured by fixed effects alone.

```{r tbl-contrasts-stream}
contrasts_stream |>
  kable(
    booktabs = TRUE, escape = TRUE, 
    caption = "Pairwise contrasts among stream effects on predicted spawn timing. Contrasts represent estimated differences in predicted spawn day of year between streams from the final model. Positive values indicate later predicted spawn timing in the first stream compared to the second. P-values are uncorrected.",
    digits = 2
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

\newpage

**Year contrasts**

There was a clear trend toward later spawning over the four-year period (Table A4.4). Spawning in 2005 occurred significantly later than in all previous years. Differences between 2002 and 2003 were not statistically significant, but later years (2004 and especially 2005) were associated with a progressive delay in mean spawn timing. This temporal shift likely reflects interannual variability in temperature and flow conditions.

```{r tbl-contrasts-year}
contrasts_year |>
  kable(
    booktabs = TRUE, escape = TRUE, 
    caption = "Pairwise contrasts among year effects on predicted spawn timing. Contrasts represent estimated differences in predicted spawn day of year between years from the final model. Positive values indicate later spawning in the first year compared to the second. P-values are uncorrected.",
    digits = 2
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

\newpage

## Reach-level variation (random effects)

```{r ranefs}
random <- estimate_grouplevel(mod_final)

random_wide <- random |>
  select(Level, Parameter, Coefficient) |>
  pivot_wider(names_from = Parameter, values_from = Coefficient) |>
  rename(COMID = Level, Intercept = `(Intercept)`, Slope = temp_90) |>
  left_join(xref_comid_stream, by = "COMID") |>
  left_join(df_elev_slope |> mutate(COMID = as.character(COMID)), by = "COMID")
```

Random intercepts and slopes varied considerably among COMIDs, with substantial spread both within and between streams (@fig-ranefs-panel A-B). Random intercepts represent reach-level deviations in baseline spawn timing after accounting for fixed effects: a positive intercept indicates a reach where salmon spawn later than expected given temperature, stream identity, and year; a negative intercept indicates earlier-than-expected spawning. Random slopes represent deviations in thermal sensitivity: steeper slopes indicate stronger temperature dependence, shallower slopes reflect weaker or buffered responses. Among-reach variance was substantial for both baseline timing (128.20) and thermal sensitivity (33.44), and the pattern of variation differed among streams.

The intercept-slope correlation was strongly negative (−0.83; @fig-ranefs-panel C), indicating that reaches with earlier baseline spawn timing exhibited substantially greater thermal sensitivity. Early-spawning, likely higher-elevation reaches are not only earlier — they are also more responsive to temperature variation, suggesting greater phenotypic plasticity. In contrast, later-spawning, lower-elevation reaches exhibit shallower thermal responses, possibly reflecting operation near thermal thresholds or stronger constraint by other cues such as photoperiod or migration timing.

@fig-ranefs-panel D translates these random effect patterns into predicted temperature–spawn timing relationships for individual COMIDs, illustrating how the combination of varying intercepts and slopes generates a diverse family of reach-specific trajectories around the population mean. While the overall relationship between pre-spawn temperature and spawn timing is positive and nonlinear, the strength and shape of this response varies substantially among reaches, likely reflecting fine-scale differences in hydrology, geomorphology, and local thermal buffering capacity. This reach-level heterogeneity is the mechanism by which fine-scale habitat diversity sustains diverse spawning portfolios within the Middle Fork.

```{r ranefs-plots}
p.re.forest <- plot(random) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme(axis.text.y = element_blank()) +
  labs(x = "Level (COMID)")

p.re.int.box <- random_wide |>
  ggplot(aes(x = stream, y = Intercept, color = stream)) +
  geom_boxplot() +
  geom_jitter(size = 2, width = 0.2, alpha = 1) +
  scale_color_manual(values = cols.streams) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(color = "none") +
  labs(x = "", y = "Random intercept\n(Baseline spawning)")

p.re.slp.box <- random_wide |>
  ggplot(aes(x = stream, y = Slope, color = stream)) +
  geom_boxplot() +
  geom_jitter(size = 2, width = 0.2, alpha = 0.5) +
  scale_color_manual(values = cols.streams) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(color = "none") +
  labs(x = "", y = "Random slope\n(Thermal responsiveness)")

rho <- as.data.frame(VarCorr(mod_final))
cor_val_model <- rho$sdcor[rho$var2 == "temp_90" & rho$var1 == "(Intercept)" & !is.na(rho$var2)]

p.re.cor <- random_wide |>
  ggplot(aes(x = Intercept, y = Slope)) +
  geom_point(aes(fill = stream), shape = 21, size = 2.5, color = "black") +
  scale_fill_manual(values = cols.streams) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  annotate(
    "text",
    x = min(random_wide$Intercept),
    y = max(random_wide$Slope),
    label = paste0("rho == ", round(cor_val_model, 2)),
    parse = TRUE,
    hjust = -5.5, vjust = 1.5
  ) +
  labs(x = "Random intercept", y = "Random slope") +
  theme(legend.position = "none")
```

```{r ranefs-plot-temp}
obs_range <- df_mod_plot |>
  group_by(COMID) |>
  summarise(
    min_temp = min(temp_90_raw),
    max_temp = max(temp_90_raw),
    .groups = "drop"
  )

me <- predict_response(
  mod_final,
  terms = c("temp_90 [all]", "COMID"),
  type = "random"
)

me_filtered <- me |>
  as_tibble() |>
  mutate(x = x * sd_temp + mean_temp) |>
  left_join(obs_range, by = c("group" = "COMID")) |>
  filter(x >= min_temp & x <= max_temp)

obs_global <- df_mod_plot |>
  summarise(
    min_temp = min(temp_90_raw),
    max_temp = max(temp_90_raw)
  )

me2 <- predict_response(
  mod_final,
  terms = c("temp_90 [all]"),
  interval = "confidence"
) |>
  as_tibble() |>
  mutate(x = x * sd_temp + mean_temp) |>
  filter(x >= obs_global$min_temp & x <= obs_global$max_temp)

p.re.preds <- me_filtered |>
  as_tibble() |>
  left_join(xref_comid_stream, by = c("group" = "COMID")) |>
  ggplot(aes(x, fixed_date + predicted, group = group)) +
  geom_point(
    data = df_mod_plot,
    aes(x = temp_90_raw, y = fixed_date + yday, color = stream),
    alpha = 0.15, size = 0.8, inherit.aes = FALSE
  ) +
  geom_line(aes(color = stream), linewidth = 0.6) +
  geom_ribbon(
    data = me2,
    aes(x = x, ymin = fixed_date + conf.low, ymax = fixed_date + conf.high),
    alpha = 0.1, inherit.aes = FALSE
  ) +
  geom_line(
    data = me2,
    aes(x = x, y = fixed_date + predicted),
    linewidth = 2, inherit.aes = FALSE
  ) +
  scale_color_manual(values = cols.streams) +
  coord_cartesian(ylim = as.Date(c("2001-07-01", "2001-09-15"))) +
  guides(color = "none") +
  labs(
    x = "90-day mean temperature pre-spawn (°C)",
    y = "Spawn date"
  )
```

```{r fig-ranefs-panel, fig.width=8, fig.height=9, out.width="90%", results='hide', fig.cap="Reach-level (COMID) random effects and predicted spawn timing relationships from the final mixed-effects model. (A) Random intercepts by stream, representing reach-level deviations in baseline spawn timing after accounting for fixed effects; positive values indicate later-than-expected spawning. (B) Random slopes by stream, representing reach-level deviations in thermal sensitivity; larger positive values indicate stronger temperature dependence. Points in A and B represent individual COMIDs. (C) Correlation between random intercepts and slopes across all COMIDs, colored by stream; the strong negative relationship (ρ = −0.83) indicates that reaches with earlier baseline spawn timing exhibit greater thermal sensitivity. (D) Predicted temperature–spawn timing relationships for individual COMIDs (colored lines) and the population mean response (black line with 95% confidence interval)."}

# (p.re.int.box + p.re.slp.box) / (p.re.cor + p.re.preds) +
#   plot_layout(guides = "collect", heights = c(.6, .8, 1)) +
#   plot_annotation(tag_levels = "A") &
#   theme(plot.margin = margin(2, 4, 1, 4))

(p.re.int.box + p.re.slp.box) / p.re.cor / p.re.preds +
  plot_layout(guides = "collect", heights = c(.5, .8, 1.2)) +
  plot_annotation(tag_levels = "A") &
  theme(plot.margin = margin(2, 4, 1, 4))

path <- here::here("plots", "Figure_06")
ggsave(
  glue::glue("{path}.pdf"),
  plot = last_plot(),
  device = cairo_pdf,
  scale = 2.75,
  width = 7,
  height = 9,
  units = "cm"
)
pdftools::pdf_convert(
  pdf = glue::glue("{path}.pdf"),
  filenames = glue::glue("{path}.png"),
  format = "png",
  dpi = 300
)
```
